---
title: Data Visualization With R - Title and Axis Labels
date: '2016-08-22'
slug: data-visualization-with-r-title-and-axis-labels
categories:
  - data-visualization
tags:
  - base
  - visualization
---

### Introduction



This is the second post of the series **Data Visualization With R**. In the previous [post](https://rsquaredacademy.github.io/post/data-visualization-with-r-introduction/), we explored the `plot()` function and observed the different types of plots it generated. In this post, we will learn how to add:

- Title
- Subtitle
- Axis Labels

to a plot and how to modify:

- Axis range

In the previous [post](https://rsquaredacademy.github.io/post/data-visualization-with-r-introduction/), we created plots which did not have any title or labels. Such plots are of no use to any one as they do not indicate what the X and Y axis represent or the primary information being communicated by the plot. The `title` and `labels` play an important part in making the plot holistic. There are two ways to add them to a plot:

- use the relevant arguments within the `plot()` function
- use the `title()` function

We will explore both the above methods one by one, and you can choose the method most convenient to you. Let us begin with the `plot()` function:



### Libraries, Code & Data



All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/7c0968878e8cf989b669ae80ac32d15a).



### Syntax



|Feature|Argument|Value|Example|
|:-------|:--------|:-----|:-------|
|Title  |main    |string|"Scatter Plot"|
|Subtitle  |sub    |string|"Displacement vs Miles Per Gallon"|
|X Axis Label  |xlab    |string|"Displacement"|
|Y Axis Label  |ylab    |string|"Miles Per Gallon"|



### Title



You can add a title to the plot using the `main` argument in the `plot()` function. Ensure that the title is enclosed in single/double quotes as it is a string. Let us create a scatter plot of `disp` and `mpg` from `mtcars` data set, and add a title to it.

```{r title, fig.align='center',fig.height=5,fig.width=5}
plot(mtcars$disp, mtcars$mpg, 
     main = 'Displacement vs Miles Per Gallon')
```



### Subtitle



You can add a subtitle to the plot using the `sub` argument in the `plot()` function. The subtitle will appear below the X axis label. Ensure that the subtitle is enclosed in single/double quotes as it is a string. Let us add a subtitle to the plot we created in the previous example:

```{r subtitle, fig.align='center',fig.height=5,fig.width=5}
plot(mtcars$disp, mtcars$mpg, 
     main = 'Displacement vs Miles Per Gallon',
     sub = 'Mileage is inversely related to Displacement')
```



### Axis Labels



In the plots created in the previous examples, the axis labels appear as `mtcars$mpg` and `mtcars$disp`. It is not the best way to name the axis and it will make more sense to use names that describe the data. Let us modify the axis labels using the `xlab` and `ylab` arguments in the `plot()` function:

```{r axis, fig.align='center',fig.height=5,fig.width=5}
plot(mtcars$disp, mtcars$mpg, 
     main = 'Displacement vs Miles Per Gallon',
     sub = 'Mileage is inversely related to Displacement',
     xlab = 'Displacement', ylab = 'Miles Per Gallon')
```



### title()



We can add title, subtitle and axis labels using the `title()` function as well. Let us recreate the plots from the previous examples but this time we will use the `title()` instead of the `plot()` function. We will continue to use the `plot()` function to create the plot.

```{r title1, fig.align='center',fig.height=5,fig.width=5, collapse=TRUE}
# create scatter plot
plot(mtcars$disp, mtcars$mpg) 

# add title, subtitle and axis labels
title(main = 'Displacement vs Miles Per Gallon',
      sub = 'Mileage is inversely related to Displacement',
      xlab = 'Displacement', ylab = 'Miles Per Gallon')
```



Do you notice that the axis labels are overwritten? This happens because the `plot()` function adds the default labels and we add a new set of labels without modifying the existing ones. The solution is to instruct the `plot()` function not to add any labels to the X and Y axis. This can be achieved using the `ann` (annotate) argument in the `plot()` function and set it to `FALSE`. Let us try it:

```{r title2, fig.align='center',fig.height=5,fig.width=5}
# create scatter plot
plot(mtcars$disp, mtcars$mpg, ann = FALSE) 

# add title, subtitle and axis labels
title(main = 'Displacement vs Miles Per Gallon',
      sub = 'Mileage is inversely related to Displacement',
      xlab = 'Displacement', ylab = 'Miles Per Gallon')
```



The axis labels are legible and not overwritten. You can use either the `plot()` function or the `title()` function to add title, subtitle and axis labels but ensure that in case you use the `title()` function, set `ann` argument to `FALSE` in the `plot()` function.




### Axis Range



In certain cases, you would want to modify the range of the axis of the plots. By default, the `plot()` function will take into account the `min` and `max` values of the variable(s) and set the range for the axis. We can modify the range by using the `xlim` and `ylim` arguments in the `plot()` function. Both the `xlim` and `ylim` arguments take 2 values as inputes. The first value is the minimum value for the axis and the second value is the maximum value for the axis. The `plot()` function will return an error if we do not specify two values for both `xlim` and `ylim` arguments. Let us recreate the plot from the previous examples but change the range of both the X and Y axis:

```{r axisrange, fig.align='center',fig.height=5,fig.width=5}
plot(mtcars$disp, mtcars$mpg, 
     xlim = c(0, 600), ylim = c(0, 50))
```

Keep in mind that the axis ranges cannot be modified using the `title()` function.



### Summary



In this post, we learned how to add title, subtitle and axis labels to a plot using the `plot()` and `title()` function. We also learned how to modify the range of the X and Y axis using the `xlim` and `ylim` arguments within the `plot()` function. 



### Up Next..



In the next [post](https://rsquaredacademy.github.io/blog/post/data-visualization-with-r-scatter-plots/), we will learn to create the **scatter plot**.

