---
title: Vectors - Part 1
date: '2016-03-03'
slug: introduction-to-vectors
categories:
  - r-introduction
tags:
  - vectors
---

## Introduction

This is the fifth post in the series **Introduction to R**. In the previous [post](), we learnt about the basic data types in R. IN this post, we will:

- understand the concept of vectors
- learn to create vectors of different data types


## Vectors

Vector is the most basic data structure in R. It is a sequence of elements of the same data type. If the elements are of different data types, they will be coerced to a common type that can accommodate all the elements. Vectors are generally created using `c()` (concatenate function), although depending on the data type of vector being created, other methods can be used.

## Numeric Vector

We will create a numeric vector using `c()` but you can use any function that creates a sequence of numbers. After that we will use `is.vector()` to check if it is a vector and `class` to check the data type.

```{r vect1}
# create a numeric vector
num_vect <- c(1, 2, 3)

# display the vector
num_vect

# check if it is a vector
is.vector(num_vect)

# check data type
class(num_vect)
```

Let us look at other ways to create a sequence of numbers. We leave it as an exercise to the reader to understand the functions using `help`.

```{r vect2}
# using colon
vect1 <- 1:10
vect1

# using rep
vect2 <- rep(1, 5)
vect2

# using seq
vect3 <- seq(10)
vect3
```

## Integer Vector

Creating an integer vector is similar to numeric vector except that we need to instruct R to treat the data as `integer` and not `numeric` or `double`. We will use the same methods we used for creating numeric vectors. To specify that the data is of type `integer`, we suffix the number with `L`.

```{r vect3}
# integer vector
int_vect <- c(1L, 2L, 3L)
int_vect

# check data type
class(int_vect)

# using colon
vect1 <- 1L:10L
vect1

# using rep
vect2 <- rep(1L, 5)
vect2

# using seq
vect3 <- seq(10L)
vect3
```

## Character Vector

A character vector may contain a single character, a word or a group of words. The elements must be enclosed in single or double quotations.

```{r vect4}
# character vector
greetings <- c("hello", "good morning")
greetings

# check data type
class(greetings)
```

## Logical Vector

A vector of logical values will either contain `TRUE` or `FALSE` or both.

```{r vect5}
# logical vector
vect_logic <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
vect_logic

# check data type
class(vect_logic)
```

In fact, you can create an `integer` vector and coerce it to type `logical`.

```{r vect6}
# integer vector
int_vect <- rep(0L:1L, 3)
int_vect

# coerce to logical vector
log_vect <- as.logical(int_vect)
log_vect

# check data type
class(log_vect)
```
