---
title: "Data Visualization with R - Legends"
date: '2016-11-25'
categories: 
  - data-visualization
tags:
  - base
  - visualization
  - legend
slug: data-visualization-with-r-legends
---

### Introduction



This is the eighth post in the series **Data Visualisation With R**. In the previous [post](https://rsquaredacademy.github.io/blog/post/data-visualization-with-r-histogram/), we learnt to build histograms. In this post, we will learn how to:

- position the legend within the plot
- modify the layout using `ncol` and `horiz` arguments
- add title using the `title.` set of arguments
- modify the appearance and position of the legend box
- modify the appearance of the text in the legend box

Legends are used to convey information about the data being represented by a plot. To understand the importance of legends, let us look at the two plots below. In the first plot, would you be able to understand what the lines represent in the absence of a legend? May be yes but only if the author provides information in a textual form outside the plot. While such information will be useful, it will also be very inconvenient. Now look at the second plot, from the legend at the top right we can easily interpret what the lines represent. Would you agree that a legend is integral to plot representing multiple data? If yes, let us go ahead and learn how to add a legend to different plots.

Since we have looked at a line graph in the above example, we will learn how to add a legend to a line graph. After that, we will generalise the steps to different plots. If you do not know how to create line plots in R, please refer to this [post](https://rsquaredacademy.github.io/blog/post/data-visualization-with-r-line-graphs/).



### Libraries, Code & Data



All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/0036a9cc1ca779965f45c38af260dcdc).



### Data



Let us build a line graph that represents annual economic growth (GDP) data of the BRICS nations for the years 2010-14.

```{r}
year <- seq(2010, 2014, 1)
india <- c(10.3, 6.6, 5.6, 6.6, 7.2)
china <- c(10.6, 9.5, 7.8, 7.7, 7.3)
russia <- c(4.5, 4.3, 3.5, 1.3, 0.7)
brazil <- c(7.5, 3.9, 1.9, 3.0, 0.1)
s_africa <- c(3.2, 3.2, 2.2, 2.2, 1.5)
gdp <- data.frame(year, india, china, russia, brazil, s_africa, stringsAsFactors = FALSE)
gdp
```



### Line Graph



Below is the line graph that represents the above GDP data set:

```{r line1, fig.align='center', fig.width= 7, fig.height=7, echo=FALSE}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
lines(gdp$year, gdp$russia, col = 'green')
lines(gdp$year, gdp$brazil, col = 'yellow')
lines(gdp$year, gdp$s_africa, col = 'violet')}
```

Without a legend, it will be very difficult to map the lines to the BRICS nations. We will add a legend to the above plot using the `legend()` function and do so one step at a time.



### Legend Location



In order to add a legend to the plot, the first thing we must specify is the location of the legend in the plot. There are 2 ways to do this:

- use x and y axis coordinates
- use keywords

The list of keywords include:

- top
- bottom
- left
- right
- center
- bottomright
- bottomleft
- topright
- topleft

But there is a glitch. If we do not specify what goes into the legend, the `legend()` function will return an error. Before we experiment with the location of the legend inside the plot, we need to learn about another argument used to specify the content of the legend. The argument is also named `legend`. It takes a vector as input. In the next example, we will plot the GDP data for India and China and add a basic legend.

```{r leg1, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2013, y = 14, legend = c('india', 'china'))}
```



You can see that a legend has been added bases on the X and Y axis coordinates we specified in the `legend()` function. But the legend is incomplete and a user still cannot map the lines to the countries using the legend. We will learn how to add lines inside the legend shortly but before that let us use keywords to position the legend inside the plot. 

```{r leg2, fig.align='center', fig.width=15, fig.height=15, echo=FALSE}
init <- par(no.readonly = TRUE)
par(mfrow = c(3, 3))
values <- c('top', 'bottom', 'left', 'right', 'center',
            'bottomright', 'bottomleft', 'topright', 'topleft')
for (i in values) {
  plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
  lines(gdp$year, gdp$china, col = 'red')
  legend(x = i, legend = c('india', 'china'))
  mtext(paste("x = ", i), side = 3, col = "blue")
}
```

You can either use the keywords or the axis coordinates to position the legend inside the plot. Use the coordinates method if you want greater control over the position of the legend. Next step is to add lines inside the legend so that a user can map the lines in the plots to the countries.



### Lines



Adding a line in the legend is very simple. Use the `lty` argument to specify the line type and the `col` argument to add color to the line. 

```{r leg3, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2012, y = 14, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'))}
```

Now we can map the lines to the respective countries using the legend. But our legend looks very simple right. Let us explore the options available to modify and enhance the appearance of the legend.



### Points



If the plot has both lines and points, we can use the `pch` argument in the `legend()` function to specify the shape of the point.

```{r leg4, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
points(gdp$year, gdp$india, pch = 2, col = 'blue')
lines(gdp$year, gdp$china, col = 'red')
points(gdp$year, gdp$china, pch = 2, col = 'red')
legend(x = 2012, y = 14, legend = c('india', 'china'),
       lty = 1, pch = 2, col = c('blue', 'red'))}
```



### Text Placement



The contents of the legend can be positioned horizontally using the `horiz` argument. It takes logical values as inputs and the default is `FALSE`. Set it to `TRUE` to position the contents horizontally:

```{r leg5, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2011, y = 14, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'),
       horiz = TRUE)}
```



Another way to position the content inside the legend is to use columns. In the below example, we use the `ncol` argument to split the contents of the legend into two columns instead of the default one. 


```{r leg6, fig.align='center', fig.width= 7, fig.height=7, echo=FALSE}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
lines(gdp$year, gdp$russia, col = 'green')
lines(gdp$year, gdp$brazil, col = 'yellow')
legend(x = 2012, y = 0, legend = c('india', 'china', 'russia', 'brazil'),
       lty = 1, col = c('blue', 'red', 'green', 'yellow'),
       ncol = 2)}
```



The below plots show the difference in appearance:

```{r leg7, fig.align='center', fig.width=15, fig.height=10, echo=FALSE}
init <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))
values <- c(1, 2, 3, 4)
for (i in values) {
  plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
  lines(gdp$year, gdp$china, col = 'red')
  lines(gdp$year, gdp$russia, col = 'green')
  lines(gdp$year, gdp$brazil, col = 'yellow')
  legend(x = 2010, y = 0, legend = c('india', 'china', 'russia','brazil'),
       lty = 1, col = c('blue', 'red', 'green', 'yellow'),
       ncol = i)
  mtext(paste("ncol = ", i), side = 3, col = "blue")
}
```



### Title



Let us add a title to the legend using the `title` argument:

```{r leg8, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2012, y = 14, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'),
       title = 'Nations')}
```



#### Title Color



The color of the title can be modified using the `title.col` argument:

```{r leg9, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2012, y = 14, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'),
       title = 'Nations', title.col = 'red')}
```



#### Title Position



The title can be positioned within the legend box using the `title.adj` argument. It will take values between 0 and 1. The default is 0.5 and the title is positioned in the middle of the box. As the value moves away from 0.5, the position of the title moves to the left or right respectively.

```{r leg10, fig.align='center', fig.height=5, fig.width=7}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2012, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), horiz = TRUE,
       title = 'Nations', title.adj = 0.1)}
```



The below plots show the relative position of the title within the legend box for different values between 0 and 1.

```{r leg11, fig.align='center', fig.width=15, fig.height=10, echo=FALSE}
init <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))
values <- c(0.2, 0.4, 0.6, 0.8)
for (i in values) {
  plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
  lines(gdp$year, gdp$china, col = 'red')
  legend(x = 2010, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), horiz = TRUE,
       title = 'Nations', title.adj = i)
  mtext(paste("title.adj = ", i), side = 3, col = "blue")
}
```



### Box Appearance



There are a lot of options to modify the appearance of the legend box. The below table displays the arguments and their descriptions. Let us look at them one by one:

```{r kable, echo=FALSE}
option <- c('Box Type', 'Background Color', 'Border Line Type',
            'Border Line Width', 'Border Line Color', 
            'Horizontal Justification', 'Vertical Justification',
            'Text Color', 'Text Font')
argument <- c('bty', 'bg', 'box.lty', 'box.lwd', 'box.col',
              'xjust', 'yjust', 'text.col', 'text.font')
values <- c('o, n', 'blue, #0000ff', '1:5', '0.5, 1, 1.5',
            'blue, #0000ff', '0:1', '0:1', 'blue, #0000ff',
            '1:5')
legend <- data.frame(option, argument, values)
knitr::kable(legend, format = "markdown", caption = "Legend Box Options")
```



#### Box Type



The `bty` argument takes two values, `o` and `n`. If set to `n`, there will be no box around the legend.

```{r leg12, fig.align='center', fig.width=10, fig.height=5, echo=FALSE}
init <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
values <- c('o', 'n')
for (i in values) {
  plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
  lines(gdp$year, gdp$china, col = 'red')
  legend(x = 2012, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), bty = i,
       title = 'Nations')
  mtext(paste("bty = ", i), side = 3, col = "blue")
}
```



### Background Color



A background color can be added to the legend box using the `bg` argument. Below is an example:

```{r leg13, fig.align='center', fig.height=5, fig.width=5}
{c}
```



### Border



The following arguments can be used to modify the border of the legend box:

- `box.lty`: line type
- `box.lwd`: line width
- `box.col`: color


```{r leg14, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2012, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), title = 'Nations',
       box.lty = 3, box.lwd = 1.5, box.col = 'green')}
```



### Justification



The `xjust` and `yjust` arguments can be used to position the legend relative to the **X** and **Y** axis respectively. Listed below is the value and the respective justification:

- `0`: left justified
- `0.5`: centered
- `1`: right justified

Let us look at a few examples to understand how it works.



#### Horizontal Justification


 
```{r leg15, fig.align='center', fig.width=15, fig.height=5, echo=FALSE}
init <- par(no.readonly = TRUE)
par(mfrow = c(1, 3))
values <- c(0, 0.5, 1)
for (i in values) {
  plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
  lines(gdp$year, gdp$china, col = 'red')
  legend(x = 2012, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), xjust = i,
       title = 'Nations')
  mtext(paste("xjust = ", i), side = 3, col = "blue")
}
```



#### Vertical Justification



```{r leg16, fig.align='center', fig.width=15, fig.height=5, echo=FALSE}
init <- par(no.readonly = TRUE)
par(mfrow = c(1, 3))
values <- c(0, 0.5, 1)
for (i in values) {
  plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
  lines(gdp$year, gdp$china, col = 'red')
  legend(x = 2012, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), yjust = i,
       title = 'Nations')
  mtext(paste("xjust = ", i), side = 3, col = "blue")
}
```



### Text Appearance



The last topic we will explore is the appearance of the text inside the legend box. We will modify the color and font of text using the `text.col` and `text.font` arguments.

```{r leg17, fig.align='center', fig.height=5, fig.width=5}
{plot(gdp$year, gdp$india, type = 'l',
      ylim = c(-5, 15), xlab = 'Year',
      ylab = 'Growth Rate', col = 'blue',
      main = 'BRICS: Growth Rate')
lines(gdp$year, gdp$china, col = 'red')
legend(x = 2012, y = 0, legend = c('india', 'china'),
       lty = 1, col = c('blue', 'red'), title = 'Nations',
       text.col = 'green', text.font = 3)}
```



### Summary



In this post, we learnt to add legend to a plot using the `legend()` function. Specifically, we learnt to

- position the legend within the plot
- modify the layout using `ncol` and `horiz` arguments
- add title using the `title.` set of arguments
- modify the appearance and position of the legend box
- modify the appearance of the text in the legend box



### Up Next..



In the next [post](https://rsquaredacademy.github.io/blog/post/data-visualization-with-r-text-annotations/), we will learn how to add text annotations to a plot using the `text()` and `mtext()` functions.









- 