---
title: Matrices - Part 2
author: ''
date: '2016-05-05'
slug: matrix-part-2
categories:
  - r-introduction
tags:
  - matrix
---

## Introduction

This is the ninth post in the series **Introduction to R**. In the previous [post](), we learnt to create matrices. In this post, we will learn to:

- combining matrices
- index/subset matrices
- dissolve matrices

## Append Data

In this section, we will learn how to append data to a matrix. There are two functions that can be used for this purpose:

- `rbind()`
- `cbind()`

`cbind` will append a new column to the matrix while `rbind` will append a new row. 

#### Append Row/Column

```{r mat20}
# 3 x 3 matrix
mat <- matrix(data = 1:9, nrow = 3)
mat

# numeric vector
v <- c(10, 11, 12)
v

# append row
rbind(mat, v)

# append column
cbind(mat, v)
```

## Combine Matrices

When you use `rbind` to combine two matrices, the number of columns must match and in case of `cbind`, the number of rows must match.

#### Append Row/Column

```{r mat21}
# 3 x 3 matrix
mat1 <- matrix(data = 1:9, nrow = 3)
mat2 <- matrix(data = sample(9), nrow = 3)

# append rows
rbind(mat1, mat2)

# append columns
cbind(mat1, mat2)
```

## Subset Matrices

In this section, we will learn to subset matrices. The `[]` operator can be used to subset matrices just like vectors but since matrices are two dimensional, we need to specify both the row number and the column number. Below are a few examples:

```{r mat22}
# 3 x 4 matrix
mat <- matrix(data = 1:12, nrow =3)
mat

# extract element from first row, first column
mat[1, 1]

# extract all rows of first column
mat[, 1]

# extract all columns of first row
mat[1,]

# extract 2nd and 3rd row of first column
mat[c(2, 3), 1]

# extract 2nd and 3rd column of first row
mat[1, c(2, 3)]

# extract 2nd and 3rd row of first and third column
mat[c(2, 3), c(1, 3)]
```

#### Using Row & Column Names

In an earlier section, we learnt how to name the rows and columns of a matrix. Let us see how these names can be used to subset matrices.

```{r mat23}
# row names
row_names <- c('row_1', 'row_2', 'row_3')

# column names
col_names <- c('col_1', 'col_2', 'col_3')

# matrix with row and column names
mat <- matrix(data = 1:9, nrow = 3, dimnames = list(row_names, col_names))

# extract elements from first row/columns
mat['row_1', 'col_1']

# extract all rows of first column
mat[, 'col_1']

# extract all columns of first row
mat['row_1',]
```

#### Using Logical Expressions

We can use logical expressions to subset matrices.

```{r mat24}
# 3 x 4 matrix
mat <- matrix(data = 1:12, nrow =3)
mat

# elements greater than 4
mat > 4

# extract elements greater than 4
mat[mat > 4]
```

## Dissolve Matrices

Till now we have learnt how to coerce a vector into matrix. Now let us learn how to coerce a matrix into a vector using:

- `c()`
- `as.vector()`

```{r mat25}
# 3 x 3 matrix
mat <- matrix(data = 1:9, nrow =3)
mat

# using c()
c(mat)

# using as.vector()
as.vector(mat)
```

