---
title: Data Wrangling with dplyr - Part 2
author: Aravind Hebali
date: '2017-12-17'
slug: data-wrangling-with-dplyr-part-2
categories:
  - data wrangling
tags:
  - dplyr
  - joins
---

## Introduction

In the previous [post](https://blog.aravindhebbali.com/2017/12/25/data-wrangling-with-dplyr-part-1/), 
we learnt about dplyr verbs and used them to compute average order value for an 
online retail company data. In this post, we will learn to combine tables using 
different `*_join` functions provided in dplyr. 

## Libraries, Code & Data

We will use the following packages: 

- [dplyr](http://dplyr.tidyverse.org/index.html) 
- [readr](http://readr.tidyverse.org/index.html)  

The data sets can be downloaded from [here](https://github.com/rsquaredacademy/datasets) 
and the codes from [here](https://gist.github.com/aravindhebbali/3e31f13a4194a8f1003034aa7d70d5ee).

```{r dplyr1, message=FALSE}
library(dplyr)
library(readr)
options(tibble.width = Inf)
```

## Case Study

For our case study, we will use two data sets. The first one, `order`, contains
details of orders placed by different customers. The second data set, `customer`
contains details of each customer. The below table displays the details stored
in each data set.

<center>
<img src = "/post/join_data.png" width = "800" height = "250">
</center>

Let us import both the data sets using `read_csv`.

### Data: Orders

```{r dplyr2, message=FALSE}
order <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/order.csv')
order
```

### Data: Customers

```{r dplyr, message=FALSE}
customer <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/customer.csv')
customer
```

We will explore the following in the case study:

- details of customers who have placed orders and their order details
- details of customers and their orders irrespective of whether a customer has 
placed orders or not
- customer details for each order
- details of customers who have placed orders
- details of customers who have not placed orders
- details of all customers and all orders

## Example Data

We will use another data set to illustrate how the different joins work. You can
view the example data sets below.

<center>
<img src = "/post/join.png" width = "800" height = "250">
</center>

## Inner Join

<center>
<img src = "/post/draw_inner.png" width = "800" height = "250">
</center>

Inner join return all rows from `Age` where there are matching values in 
`Height`, and all columns from `Age` and `Height`. If there are multiple matches 
between `Age` and `Height`, all combination of the matches are returned.

#### Case Study

###### Details of customers who have placed orders and their order details

To get data for all those customers who have placed orders in the past let us 
join the `order` data with the `customer` data using `inner_join`. 

```{r dplyr3}
customer %>%
  inner_join(order)
```

## Left Join

<center>
<img src = "/post/draw_left.png" width = "800" height = "250">
</center>

Left join return all rows from `Age`, and all columns from `Age` and `Height`. 
Rows in `Age` with  no match in `Height` will have NA values in the new columns. 
If there are multiple  matches between `Age` and `Height`, all combinations of 
the matches are returned.

#### Case Study

###### Details of customers and their orders irrespective of whether a customer has 
placed orders or not.

To get data for all those customers and their orders irrespective of whether a 
customer has placed orders or not let us join the `order` data with the 
`customer` data using `left_join`. 


```{r dplyr4}
customer %>%
  left_join(order)
```

## Right Join

<center>
<img src = "/post/draw_right.png" width = "800" height = "250">
</center>

Right join return all rows from `Height`, and all columns from `Age` and 
`Height`. Rows in `Height` with no match in `Age` will have NA values in the 
new columns. If there are multiple matches between `Age` and `Height`, all 
combinations of the matches are returned.

#### Case Study

###### Customer details for each order

To get customer data for all orders, let us join the `order` data with the 
`customer` data using `right_join`. 

```{r dplyr5}
customer %>%
  right_join(order)
```

## Semi Join

<center>
<img src = "/post/draw_semi.png" width = "800" height = "250">
</center>

Semi join return all rows from `Age` where there are matching values in `Height`, 
keeping just columns from `Age`. A semi join differs from an inner join because 
an inner join will return one row of `Age` for each matching row of `Height`, 
where a semi join will never duplicate rows of `Age`.

#### Case Study

###### Details of customers who have placed orders

To get customer data for all orders, let us join the `order` data with the 
`customer` data using `semi_join`. 

```{r dplyr6}
customer %>%
  semi_join(order)
```

## Anti Join

<center>
<img src = "/post/draw_anti.png" width = "800" height = "250">
</center>

Anti join return all rows from `Age` where there are not matching values in 
`Height`, keeping just columns from `Age`.

#### Case Study

###### Details of customers who have not placed orders

To get details of customers who have not placed orders, let us join the `order` 
data with the `customer` data using `anti_join`. 

```{r dplyr7}
customer %>%
  anti_join(order)
```

## Full Join

<center>
<img src = "/post/draw_full.png" width = "800" height = "250">
</center>

Full join return all rows and all columns from both `Age` and `Height`. Where 
there are not matching values, returns NA for the one missing.

#### Case Study

###### Details of all customers and all orders

To get details of all customers and all orders, let us join the `order` 
data with the `customer` data using `full_join`. 

```{r dplyr8}
customer %>%
  full_join(order)
```