---
title: Readable Code with Pipes
author: Aravind Hebbali
date: '2017-12-03'
slug: readable-code-with-pipes
categories:
  - data wrangling
tags:
  - pipes
  - magrittr
---

## Introduction

R code contain a lot of parentheses in case of a sequence of multiple operations. When you are dealing with 
complex code, it results in nested function calls which are hard to read and maintain. The [magrittr](https://CRAN.R-project.org/package=magrittr) package by [Stefan Milton Bache](http://stefanbache.dk/) provides pipes enabling us to write R code that is readable.

Pipes allow us to clearly express a sequence of multiple operations by:

- structuring operations from left to right
- avoiding
    - nested function calls
    - intermediate steps
    - overwriting of original data
- minimizing creation of local variables

## Pipes

If you are using [tidyverse](https://www.tidyverse.org/), magrittr will be automatically loaded. We will look at 3 different types of pipes:

- `%>%` : pipe a value forward into an expression or function call
- `%<>%`: result assigned to left hand side object instead of returning it 
- `%$%` : expose names within left hand side objects to right hand side expressions

## Libraries, Code & Data

We will use the following packages in this post:

- [magrittr](http://magrittr.tidyverse.org/) 
- [readr](http://readr.tidyverse.org/) 
- [dplyr](http://dplyr.tidyverse.org/)
- [stringr](http://stringr.tidyverse.org/)
- and [purrr](http://readr.tidyverse.org/) 

You can find the data sets [here](https://github.com/rsquaredacademy/datasets) and the codes 
[here](https://gist.github.com/aravindhebbali/26d85ab4a4dadd2fe7c1f58d854cc950).


```{r importlib1, message=FALSE}
library(magrittr)
library(readr)
library(dplyr)
library(stringr)
library(purrr)
```

## Data

```{r show, message=FALSE}
ecom <- read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
ecom
```

#### Data Dictionary

- id: row id
- referrer: referrer website/search engine
- os: operating system
- browser: browser
- device: device used to visit the website
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- repeat: frequency of visits
- country: country of origin
- purchase: whether visitor purchased
- order_value: order value of visitor (in dollars)

## First Example

Let us start with a simple example. You must be aware of `head()`. If not,
do not worry. It returns the first few observations/rows of data. We can
specify the number of observations it should return as well. Let us use 
it to view the first 10 rows of our data set.

```{r mag1}
head(ecom, 10)
```

#### Using Pipe

Now let us do the same but with `%>%`. 

```{r mag2}
ecom %>% head(10)
```

## Square Root 

Time to try a slightly  more challenging example. We want the square root of 
`n_pages` column from the data set. To ensure the output does not clutter 
the page, we will view the first few observations using `head()`.

```{r mag3}
y <- head(sqrt(ecom$n_pages))
```

Let us break down the above computation into small steps:

- select/expose the `n_pages` column from `ecom` data
- compute the square root
- assign the first few observations to `y`

Let us reproduce `y` using pipes.


```{r mag4}
# select n_pages variable and assign it to y
y <-
    ecom %$%
    n_pages

# compute square root of y and assign it to y 
y %<>% sqrt

head(y)
```

Another way to compute the square root of y is shown below.

```{r mag5}
y <-
  ecom %$% 
  n_pages %>% 
  sqrt 

head(y)
```

## Visualization

Let us look at a data visualization example. We will create a bar plot to
visualize the frequency of different referrer types that drove purchasers
to the website. Let us look at the steps involved in creating the bar plot:

- extract rows where purchase == true
- select/expose `referrer` column
- tabulate referrer data using `table()`
- use the tabulated data to create bar plot using `barplot()`

```{r mag21, fig.align='center', fig.height=4, fig.width=6}
barplot(table(subset(ecom, purchase == 'true')$referrer))
```

#### Using pipe

<center>
<img src = "/post/pipes_viz.png" width = "800" height = "600">
</center>


```{r mag7, fig.align='center', fig.height=4, fig.width=6}
ecom %>%
  subset(purchase == 'true') %>%
  extract('referrer') %>%
  table %>%
  barplot
```

## Correlation

Correlation is a statistical measure that indicates the extent to which two or more variables 
fluctuate together. In R, correlation is computed using `cor()`. Let us look at the 
correlation between the number of pages browsed and time spent on the site for
visitors who purchased some product. Below are the steps for computing correlation:

- extract rows where purchase == true
- select/expose `n_pages` and `duration` columns
- use `cor()` to compute the correlation

```{r mag6, collpase = TRUE}
# without pipe
ecom1 <- subset(ecom, purchase == 'true')
cor(ecom1$n_pages, ecom1$duration)

# with pipe
ecom %>%
  subset(purchase == 'true') %$% 
  cor(n_pages, duration)

# with pipe
ecom %>%
  filter(purchase == 'true') %$% 
  cor(n_pages, duration)
```


## Regression

Let us look at a regression example. We regress time spent on the site on 
number of pages visited. Below are the steps involved in running the regression:

- use `duration` and `n_pages` columns from econ data
- pass the above data to `lm()`
- pass the output from `lm()` to `summary()` 

```{r mag8}
summary(lm(duration ~ n_pages, data = ecom))
```

#### Using pipe

```{r mag22}
ecom %$%
  lm(duration ~ n_pages) %>%
  summary
```

## String Manipulation

We want to extract the first name (jovial) from the below email id and
convert it to upper case. Below are the steps to achieve this:

- split the email id using the pattern `@` using `str_split()`
- extract the first element from the resulting list using `extract2()`
- extract the first element from the character vector using `extract()`
- extract the first six characters using `str_sub()`
- convert to upper case using `str_to_upper()`

```{r mag9}
email <- 'jovialcann@anymail.com'

# without pipe
str_to_upper(str_sub(str_split(email, '@')[[1]][1], start = 1, end = 6))

# with pipe
email %>%
  str_split(pattern = '@') %>%
  extract2(1) %>%
  extract(1) %>%
  str_sub(start = 1, end = 6) %>%
  str_to_upper()
```

<center>
<img src = "/post/pipes_canva.png" width = "800" height = "600">
</center>

Another method that uses `map_chr()` from the [purrr](http://readr.tidyverse.org/) package.

```{r map39}
email %>%
  str_split(pattern = '@') %>%
  map_chr(1) %>%
  str_sub(start = 1, end = 6) %>%
  str_to_upper()
```

## Data Extraction

Let us turn our attention towards data extraction. magrittr provides
alternatives to `$`, `[` and `[[`.

- `extract()`
- `extract2()`
- `use_series()`

#### Extract Column By Name

To extract a specific column using the column name, we mention the name 
of the column in single/double quotes within `[` or `[[`. In case of `$`,
we do not use quotes.

Let us extract the first 3 observations of `n_pages` column.

```{r mag10}
head(ecom['n_pages'], 3)

ecom %>%
  extract('n_pages') %>%
  head(3)
```

#### Extract Column By Position

We can extract columns using their index position. Keep in mind that index 
position starts from **1** in R. In the below example, we show how to 
extract the first 3 observations of `n_pages` column but instead of using
the column name, we use the column position.

```{r mag23}
head(ecom[6], 3)

ecom %>%
  extract(6) %>%
  head(3)
```

#### Extract Column (as vector)

One important differentiator between `[` and `[[` is that `[[` will
return a atomic vector and not a `data.frame`. `$` will also return
a atomic vector. In magrittr, we can use `use_series()` in place of
`$`.

```{r mag11}
head(ecom$n_pages)

ecom %>%
  use_series('n_pages') %>%
  head
```

#### Extract List Element By Name

To extract elements of a list, we can use `extract2()`. If is an
alternative for `[[`.

```{r mag12}
mt <- as.list(mtcars)

mt[['mpg']]

mt %>%
  extract2('mpg')
```

#### Extract List Element By Position

```{r mag13}
mt <- as.list(mtcars)

mt[[1]]

mt %>%
  extract2(1)
```

#### Extract List Element 

We can extract the elements of a list using `use_series()` as well.

```{r mag14}
mt <- as.list(mtcars)

mt$mpg

mt %>%
  use_series(mpg)
```

## Arithmetic Operations

magrittr offer alternatives for arithemtic operations as well. We will look at 
a few examples below.

- `add()`
- `subtract()`
- `multiply_by()`
- `multiply_by_matrix()`
- `divide_by()`
- `divide_by_int()`
- `mod()`
- `raise_to_power()`

#### Addition

```{r mag15}
1:10 %>%
  `+`(1)

1:10 %>%
  add(1)
```

#### Multiplication

```{r mag16}
1:10 %>%
  `*`(3)

1:10 %>%
  multiply_by(3)
```

#### Division

```{r mag17}
1:10 %>%
  `/`(2)

1:10 %>%
  divide_by(2)
```

#### Power

```{r mag18}
1:10 %>%
  `^`(2)

1:10 %>%
  raise_to_power(2)
```

## Logical Operators

There are alternatives for logical operators as well. We will look at 
a few examples below.

- `and()`
- `or()`
- `equals()`
- `not()`
- `is_greater_than()`
- `is_weakly_greater_than()`
- `is_less_than()`
- `is_weakly_less_than()`

#### Greater Than

```{r mag19}
1:10 %>%
  `>`(5)

1:10 %>%
  is_greater_than(5)
```

#### Weakly Greater Than

```{r mag20}
1:10 %>%
  `>=`(5)

1:10 %>%
  is_weakly_greater_than(5)
```
