---
title: 'ggplot2 - Introduction to geoms'
author: Aravind Hebbali
twitterImg: /img/gg_geoms.png
description: "ggplot2 - Introduction to geoms"
date: '2017-10-15'
slug: ggplot2-introduction-to-geoms
topics:
  - ggplot2
tags:
  - ggplot2
  - visualization
---

### Introduction

This is the third post in the series **Elegant Data Visualization with 
ggplot2**. In the previous post, we learnt how to create plots using the 
`qplot()` function. In this post, we will create some of the most routinely 
used plots to explore data using the `geom_*` functions. 

```{r geoms1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
```

<br>

### Libraries, Code & Data

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)
- [tibble](http://tibble.tidyverse.org/)
- [dplyr](http://dplyr.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/2d0d5d5b60d0ef0f4d1b227c8fb0335d).

<br>

### Data

```{r use, eval=TRUE, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
ecom
```

<br>

#### Data Dictionary

- id: row id
- referrer: referrer website/search engine
- os: operating system
- browser: browser
- device: device used to visit the website
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- repeat: frequency of visits
- country: country of origin
- purchase: whether visitor purchased
- order_value: order value of visitor (in dollars)

<br>

### Scatter Plot

A scatter plot displays the relationship between two continuous variables. In 
ggplot2, we can build a scatter plot using `geom_point()`. Scatterplots can 
show you visually 

- the strength of the relationship between the variables 
- the direction of the relationship between the variables 
- and whether outliers exist

<br>

#### Point

The variables representing the X and Y axis can be specified either in `ggplot()`
or in `geom_point()`. We will learn to modify the appearance of the points in a 
different post.

```{r geoms2, fig.align='center'}
ggplot(ecom, aes(x = n_pages, y = duration)) + 
  geom_point()
```

<br>

#### Regression Line

A regression line can be fit using either:

- `geom_abline()`
- `geom_smooth()`

#### Regression Line

If you are using `geom_abline()`, you need to specify the intercept and slope 
as shown in the below example:

```{r geoms5, fig.align='center'}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() + 
  geom_abline(intercept = 37.285, slope = -5.344)
```

<br>

#### Regression Line

If you are using `geom_smooth()`, you need to specify the method of fitting the
line, which can be `lm` or `loess`. You also need to indicate whether the 
confidence interval must be displayed using the `se` argument.

```{r geoms6, fig.align='center'}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_smooth(method = 'lm', se = TRUE)
```

<br>

#### Loess Method

Here we use the `'loess'` method to fit the regression line.

```{r geoms7, fig.align='center'}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_smooth(method = 'loess', se = FALSE)
```

<br>

#### Horizontal/Vertical Lines

Add horizontal or vertical lines using

- `geom_hline()`
- `geom_vline()`

<br>

#### Horizontal Line

To add a horizontal line, the Y axis intercept must be supplied using the 
`yintercept` argument.

```{r geoms4, fig.align='center'}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_hline(yintercept = 30) 
```

<br>

#### Vertical Line

For the vertical line, the X axis intercept must be supplied using the 
`xintercept` argument.

```{r geoms3, fig.align='center'}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_vline(xintercept = 5) 
```

<br>

#### Bar Plot

Bar plots present grouped data with rectangular bars. The bars may represent 
the frequency of the groups or values. Bar plots can be: 

- horizontal 
- vertical
- grouped
- stacked 
- proportional

<br>

#### Frequency

```{r geoms8, fig.align='center'}
ggplot(ecom, aes(x = factor(device))) +
  geom_bar()
```

<br>

#### Weight

If the bars should represent a continuous variable, use the `weight` argument
within `aes()`. In the below example, the bars do not represent the count of 
devices, instead, they represent the total order value for each device type.

```{r geoms9, fig.align='center'}
ggplot(ecom, aes(x = factor(device))) +
  geom_bar(aes(weight = order_value))
```

<br>

#### Stacked Bar Plot

To create a stacked bar plot, the `fill` argument must be mapped to a 
categorical variable. 

```{r geoms10, fig.align='center'}
ggplot(ecom, aes(x = factor(device))) +
  geom_bar(aes(fill = purchase))
```

<br>

#### Horizontal Bar Plot

A horizontal bar plot can be created by flipping the coordinate axes using the
`coord_flip()` function.

```{r geoms11, fig.align='center'}
ggplot(ecom, aes(x = factor(device))) +
  geom_bar(aes(fill = purchase)) +
  coord_flip()
```

<br>

#### Columns

If the data has already been summarized, you can use `geom_col()` instead of 
`geom_bar()`. In the below example, we have the total visits for each device
type. The data has already been summarized and as such we cannot use `geom_bar()`.

```{r geoms12, fig.align='center', fig.height=4}
device <- c('laptop', 'mobile', 'tablet')
visits <- c(30000, 12000, 5000)
traffic <- tibble::tibble(device, visits)
ggplot(traffic, aes(x = device, y = visits)) +
  geom_col(fill = 'blue') 
```

<br>

#### Boxplot

The box plot is a standardized way of displaying the distribution of data 
based on the five number summary: minimum, first quartile, median, third 
quartile, and maximum. Box plots are useful for detecting outliers and for 
comparing distributions. It shows the shape, central tendancy and variability 
of the data. Use `geom_boxplot()` to create a box plot.

<br>

```{r geoms13, fig.align='center'}
ggplot(ecom, aes(x = factor(device), y = n_pages)) +
  geom_boxplot()
```

<br>

#### Histogram

A histogram is a plot that can be used to examine the shape and spread of 
continuous data. It looks very similar to a bar graph and can be used to detect 
outliers and skewness in data. Use `geom_histogram()` to create a histogram.

<br>

```{r geoms15, fig.align='center'}
ggplot(ecom, aes(x = duration)) +
  geom_histogram()
```

<br>

You can control the number of bins using the `bins` argument.

```{r geoms16, fig.align='center'}
ggplot(ecom, aes(x = duration)) +
  geom_histogram(bins = 5)
```

<br>

#### Line

Line charts are used to examine trends over time. We will use a different data 
set for exploring line plots.

<br>

#### Data

```{r import2, eval=TRUE, message=FALSE}
gdp <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/gdp.csv')
gdp
```

<br>

Use `geom_line()` to create a line chart. In the below plot, we chart the GDP 
of India, the fastest growing economy in emerging markets, across years.

```{r geomline1, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line()
```

<br>

The color and line type can be modified using the `color` and `linetype` 
arguments. We will explore the different line types in an upcoming post.

```{r geomline2, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line(color = 'blue', linetype = 'dashed')
```

<br>

#### Label

You can label the points using `geom_label()`.

```{r geoms20, fig.align='center'}
ggplot(mtcars, aes(disp, mpg, label = rownames(mtcars))) +
  geom_label()
```

<br>

#### Text

`geom_text()` offers another way to add text to the plots. We will learn to 
modify the appearance and location of the text in another post.

```{r geoms19, fig.align='center'}
ggplot(mtcars, aes(disp, mpg, label = rownames(mtcars))) +
  geom_text(check_overlap = TRUE, size = 2)
```

<br>

### Summary 

In this post, we learnt about different `geoms` such as 

- `geom_point()`
- `geom_line()`
- `geom_histogram()`
- `geom_bar()`
- `geom_boxplot()`
- `geom_abline()`
- `geom_text()`

<br>

### Up Next..

In the next post, we will learn about aesthetics.
