---
title: Legend - Part 6
author: ''
date: '2017-09-20'
slug: legend-part-6
topics:
  - data-visualization
tags:
  - ggplot2
  - legend
---

### Introduction

<hr>

This is the 18th post in the series **Elegant Data Visualization with ggplot2**. In the previous [post](https://rsquaredacademy.github.io/blog/post/legend-part-5/), we learnt how to modify the legend of plot when `alpha` is mapped to a categorical variable. In this post, we will learn to modify legend

- title
- label
-  and bar

<br>

```{r load, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

<br>

### Libraries, Code & Data

<hr>

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/a77a836e7e06c61c28ba7db3a69e1510).

<br>

### Title

<hr>

- position
    - top
    - bottom
    - left
    - right
- alignment
    - 0 (left)
    - 1 (right)


<br>

### Legend Title

<hr>

```{r leg1, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders"))
```

<br>

### Title Alignment

<hr>

```{r leg2, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5))
```

<br>

### Title Position

<hr>

```{r leg3, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top"))
```

<br>

### Label

<hr>

- position
    - top
    - bottom
    - left 
    - right
- alignment 
    - 0 (left)
    - 1 (right)
- direction
    - horizontal
    - vertical
- reverse
    - TRUE
    - FALSE
- rows

<br>

### Label Position

<hr>

```{r leg4, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.position = "right"))
```

<br>

### Label Alignment

<hr>

```{r leg5, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.hjust = 0.5))
```

<br>

### Direction

<hr>

```{r leg6, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(direction = "horizontal"))
```

<br>

### Rows

<hr>

```{r leg7, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(nrow = 2))
```

<br>

### Reverse

<hr>

```{r leg8, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(reverse = TRUE))
```

<br>

### Guides

<hr>

```{r leg20, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top", label.position = "right", 
      direction = "horizontal", label.hjust = 0.5, nrow = 2, reverse = TRUE)
  )
```

<br>

### Title Alignment (Vertical)

<hr>

```{r leg22, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    title = "Horsepower", title.position = "top", title.vjust = 1))
```

<br>

### Labels Alignment (Vertical)

<hr>

```{r leg23, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    label.vjust = 0.8))
```

<br>

### Legend Bar 

<hr>

- width
- height
- bins
- ticks
- limits

<br>

### Legend Bar Width

<hr>

```{r leg24, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barwidth = 10))
```

<br>

### Legend Bar Height

<hr>

```{r leg25, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barheight = 3))
```

<br>

### Legend Bar Bins

<hr>

```{r leg26, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    nbin = 4))
```

<br>

### Legend Bar Ticks

<hr>

```{r leg28, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    ticks = FALSE))
```

<br>

### Legend Bar Upper/Lower Limits

<hr>

```{r leg29, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    draw.ulim = TRUE, draw.llim = FALSE))
```

<br>

### Guides

<hr>

```{r leg34, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, 
    size = qsec, shape = factor(gear))) + 
  guides(color = "colorbar", shape = "legend", size = "legend")
```

<br>

### Guides: Color, Shape & Size

<hr>

```{r leg35, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, size = wt, shape = factor(gear))) + 
  guides(color = guide_colorbar(title = "Horsepower"),
    shape = guide_legend(title = "Weight"), size = guide_legend(title = "Gear")
  )
```

<br>

### Summary

<hr>

<br>

### Up Next..

<hr>

In the next [post](https://rsquaredacademy.github.io/blog/post/overlapping/), we will learn how to position the geoms to avoid overlapping. 

<br>