---
title: Introduction to tibbles
author: Aravind Hebbali
date: '2017-11-26'
slug: introduction-to-tibbles
categories:
  - data wrangling
tags:
  - tibble
---

## Introduction

> A **tibble**, or `tbl_df`, is a modern reimagining of the data.frame, keeping what time has proven to be effective, and throwing out what is not. Tibbles are data.frames that are lazy and surly: they do less (i.e. they don’t change variable names or types, and don’t do partial matching) and complain more (e.g. when a variable does not exist). This forces you to confront problems earlier, typically leading to cleaner, more expressive code. Tibbles also have an enhanced `print method()` which makes them easier to use with large datasets containing complex objects.

> Source: http://tibble.tidyverse.org/

In this post, we will explore tibbles. To be more precise, we will learn:

- how tibbles are different from data frames?
- how to create tibbles?
- how to manipulate tibbles?

## Libraries, Code & Data

We will use the following packages: 

- [tibble](http://tibble.tidyverse.org/) 
- [dplyr](http://dplyr.tidyverse.org/) 

The code can be found [here](https://gist.github.com/aravindhebbali/9a3814b9b4bb5c271d030b15ce4ecdf1).

```{r tibble1, message=FALSE}
library(tibble)
library(dplyr)
```

## Creating tibbles

tibble can be created using any of the following:

- `tibble()`
- `as_tibble()`
- `tribble()`

Let us start with `tibble()`. 

```{r tibble2}
tibble(x = letters,
       y = 1:26,
       z = sample(100, 26))
```

We mentioned the column names followed by the data. If you do not specify the
column names, `tibble()` will supply them. Ensure that the length of each column
is same.

## tibble features

- never changes input's types

`tibble()` will never alter the input's type. For example, if you supply a 
character vector it will not be converted to factor unlike data.frame where
you need to `stringsAsFactors` to `FALSE`.

```{r tibble3}
tibble(x = letters,
       y = 1:26,
       z = sample(100, 26))
```

- never adjusts variable names

`tibble()` will never modify the column names. In the below example, you can 
observe that while `data.frame` adds a `.`, `tibble()` retains the column names
as is.

```{r tibble4}
names(data.frame(`order value` = 10))
names(tibble(`order value` = 10))
```

- never prints all rows

`tibble()` will never print all the rows and clutter your console. It will only 
print the first 10 rows and only as many columns that fit the width of the 
console.

```{r tibble5}
x <- 1:100
y <- letters[1]
z <- sample(c(TRUE, FALSE), 100, replace = TRUE)
tibble(x, y, z)
```

- never recycles vector of length greater than 1

Recycling vectors of length greater than 1 often leads to errors and as such 
`tibble()` will only recycle vectors of length 1.

```{r tibble6, eval=FALSE}
x <- 1:100
y <- letters
z <- sample(c(TRUE, FALSE), 100, replace = TRUE)
tibble(x, y, z)
Error in overscope_eval_next(overscope, expr) : object 'y' not found
```

## Membership Testing

We can test if an object is a tibble using `is_tibble()`.

```{r tibble14}
is_tibble(mtcars)
is_tibble(as_tibble(mtcars))
```

## Tribble

Another way to create tibbles is using `tribble()`: 

- it is short for transposed tibbles
- it is customized for data entry in code 
- column names start with `~` 
- and values are separated by commas

```{r tibble15}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  1, TRUE, 'a',
  2, FALSE, 'b'
)
```

## Column Names

Names of the columns in tibbles need not be valid R variable names. They can 
contain unusual characters like a space or a smiley but must be enclosed in 
ticks. 

```{r tibble16}
tibble(
  ` ` = 'space',
  `2` = 'integer',
  `:)` = 'smiley'
)
```

## Add Rows

Let us add data related to **Safari** browser to the web traffic data using 
`add_row()`.

```{r tibble21}
browsers <- enframe(c(chrome = 40, firefox = 20, edge = 30))
browsers
```

```{r tibble22}
add_row(browsers, name = 'safari', value = 10)
```

If we want to add the data at a particular row, we can specify the row number 
using the `.before` argument. Let us add the data related to **Safari** browser 
in the second row instead of the last row.

```{r tibble23}
add_row(browsers, name = 'safari', value = 10, .before = 2)
```

## Add Columns

`add_column()` adds a new column to tibbles.

```{r tibble24}
browsers <- enframe(c(chrome = 40, firefox = 20, edge = 30, safari = 10))
add_column(browsers, visits = c(4000, 2000, 3000, 1000))
```

## Rownames

The [tibble](tibble.tidyverse.org) package provides a set of functions to deal 
with rownames. Remember, `tibble` does not have `rownames` unlike `data.frame`. 
To check whether a data set has rownames, use `has_rownames()`.

```{r tibble25}
has_rownames(mtcars)
```

#### Remove Rownames

```{r tibble26}
remove_rownames(mtcars)
```

#### Rownames to Column

```{r tibble27}
head(rownames_to_column(mtcars))
```

#### Column to Rownames

To convert the first column in the data set to rownames, use `column_to_rownames()`:

```{r tibble28}
mtcars_tbl <- rownames_to_column(mtcars)
column_to_rownames(mtcars_tbl)
```

## Glimpse

Use `glimpse()` to get an overview of the data.

```{r tibble 29}
glimpse(mtcars)
```

## Check Column 

`has_name()` can be used to check if a tibble has a specific column.

```{r tibble30}
has_name(mtcars, 'cyl')
has_name(mtcars, 'gears')
```

## Summary

#### Creating tibbles

- use `tibble()` to create tibbles
- use `as_tibble()` to coerce other objects to tibble
- use `enframe()` to coerce vector to tibble
- use `tribble()` to create tibble using data entry

#### Modifying tibbles

- use `add_row()` to add a new row
- use `add_column()` to add a new column
- use `remove_rownames()` to remove rownames from data
- use `rownames_to_colum()` to coerce rowname to first column
- use `column_to_rownames()` to coerce first column to rownames

#### Testing tibbles

- use `is_tibble()` to test if an object is a tibble
- use `has_rownames()` to check whether a data set has rownames
- use `has_name()` to check if tibble has a specific column 
- use `glimpse()` to get an overview of data
