---
title: Data Wrangling with dplyr - Part 1
author: Aravind Hebbali
date: '2017-12-10'
slug: data-wrangling-with-dplyr-part-1
categories:
  - data wrangling
tags:
  - dplyr
---

## Introduction

According to  a [survey](http://visit.crowdflower.com/rs/416-ZBE-142/images/CrowdFlower_DataScienceReport_2016.pdf) by [CrowdFlower](https://www.crowdflower.com/), data scientists spend most of their 
time cleaning and manipulating data rather than mining or modeling them for 
insights. As such, it becomes important to have tools that make data 
manipulation faster and easier. In today's post, we introduce you to 
[dplyr](http://dplyr.tidyverse.org/), a grammar of data manipulation.

<center>
<img src = "/post/crowd_flower3.png" width = "800" height = "600">
</center>


## Libraries, Code & Data

We will use the following libraries:

- [dplyr](http://dplyr.tidyverse.org/index.html) 
- and [readr](http://readr.tidyverse.org/index.html)  

The data sets can be downloaded from [here](https://github.com/rsquaredacademy/datasets) 
and the codes from [here](https://gist.github.com/aravindhebbali/7758b86c2bc13ff1e5d88d9d1c204f8c).

```{r libs, message=FALSE}
library(dplyr)
library(readr)
```

## dplyr Verbs

dplyr provides a set of verbs that help us solve the most common data 
manipulation challenges while working with tabular data (dataframes, tibbles):

- `select`: returns subset of columns
- `filter`: returns a subset of rows
- `arrange`: re-order or arrange rows according to single/multiple variables
- `mutate`: create new columns from existing columns
- `summarise`: reduce data to a single summary

## Case Study

We will explore a dummy data set that resembles web logs of an online retail 
company. You can download the data from [here](https://github.com/rsquaredacademy/datasets/blob/master/web.csv) 
or import it directly using `read_csv()` from the [readr](readr.tidyverse.org) 
package. We will use dplyr to answer the following:

- what is the average order value by device types?
- what is the average number of pages visited by purchasers and non-purchasers?
- what is the average time on site for purchasers vs non-purchasers?
- what is the average number of pages visited by purchasers and non-purchasers 
using mobile?

## Data

```{r show, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
ecom
```

###### Data Dictionary

Below is the description of the data set:

- id: row id
- referrer: referrer website/search engine
- os: operating system
- browser: browser
- device: device used to visit the website
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- repeat: frequency of visits
- country: country of origin
- purchase: whether visitor purchased
- order_value: order value of visitor (in dollars)

## Average Order Value

<center>
<img src = "/post/image.jpg" width = "800" height = "600">
</center>

## AOV by Devices

<center>
<img src = "/post/aov_flow_canva.png" width = "800" height = "600">
</center>


```{r case}
ecom %>%
  filter(purchase == 'true') %>%
  select(device, order_value, order_items) %>%
  group_by(device) %>%
  summarise_all(funs(sum)) %>%
  mutate(
    aov = order_value / order_items
  ) %>%
  select(device, aov)
```

## Syntax

Before we start exploring the dplyr verbs, let us look at their syntax:

- the first argument is always a `data.frame` or `tibble`
- the subsequent arguments provide the information required for the verbs to 
take action
- the name of columns in the data need not be surrounded by quotes


## Filter Rows

In order to compute the AOV, we must first separate the purchasers from 
non-purchasers. We will do this by filtering the data related to purchasers
using the `filter()` function. It allows us to filter rows that meet a specific 
criteria/condition. The first argument is the name of the data frame and the 
rest of the arguments are expressions for filtering the data. Let us look at a 
few examples:

The first example we will look at filters all visits from device **mobile**. As 
we had learnt in the previous section, the first argument is our data set `ecom` 
and the next argument is the condition for filtering rows.

```{r dplyr2}
filter(ecom, device == "mobile")
```

We can specify multiple filtering conditions as well. In the below example, we
specify two filter conditions:

- visit from device **tablet**
- resulted in a purchase or conversion

```{r dplyr3}
filter(ecom, device == "tablet", purchase == "true")
```

Here is another example where we specify multiple conditions: 

- visit from device **tablet**
- made a purchase
- browsed less than 15 pages 

```{r dplyr4}
filter(ecom, device == "tablet", purchase == "true", n_pages < 15)
```

##### Case Study

Let us apply what we have learnt to the case study. We want to filter all
visits that resulted in a purchase. 

```{r dplyr5}
filter(ecom, purchase == "true")
```

## Select Columns

After filtering the data, we need to select relevent variables to compute the 
AOV. Remember, we do not need all the columns in the data to compute a required 
metric (in our case, AOV). The `select()` function allows us to select a subset 
of columns. The first argument is the name of the data frame and the subsequent 
arguments specify the columns by name or position. 

To select the `device` and `purchase` columns, we specify the data set i.e. 
`ecom` followed by the name of the columns.

```{r dplyr6}
select(ecom, device, purchase)
```

We can select a set of columns using `:`. In the below example, we select all 
the columns starting from `device` up to `purchase`. Remember that we can use `:` only when
the columns are adjacent to each other in the data set.

```{r dplyr7}
select(ecom, device:purchase)
```

What if you want to select all columns except a few? Typing the name of many
columns can be cumbersome and may also result in spelling errors. We may use `:`
only if the columns are adjacent to each other but that may not always be
the case. dplyr allows us to specify columns that need not be
selected using `-`. In the below example, we select all columns except `id` and 
`country`. Notice the `-` before both of them.

```{r dplyr8}
select(ecom, -id, -country)
```

##### Case Study

For our case study, we need to select the columns `order_value` and `order_items` 
to calculate the AOV. We also need to select the `device` column as we are 
computing the AOV for each device type.

```{r dplyr9}
select(ecom, device, order_value, order_items)
```

But we want the above data only for purchasers. Let us combine `filter()` and 
`select()` functions to extract `order_value` and `order_items` only for those
visis that resulted in a purchase.

```{r dplyr10}
# filter all visits that resulted in a purchase
ecom1 <- filter(ecom, purchase == "true")

# select the relevant columns
ecom2 <- select(ecom1, device, order_value, order_items)

# view data
ecom2
```

## Grouping Data

We need to compute the total order value and total order items for each device 
in order to compute their AOV. To achieve this, we need to group the selected 
`order_value` and `order_items` by device type. `group_by()` allows us to group 
or split data based on particular (discrete) variable. The first argument is the 
name of the data set and the second argument is the name of the column based on 
which the data will be split. 

To split the data by referrer type, we use `group_by` and specify the data set
i.e. `ecom` and the column based on which to split the data i.e. `referrer`.

```{r dplyr11}
group_by(ecom, referrer)
```

##### Case Study

In the second line in the previous output, you can observe `Groups: referrer [5]`
. The data is split into 5 groups as the referrer variable has 5 distinct values. 
For our case study, we need to group the data by `device` type.

```{r dplyr12}
# split ecom2 by device type
ecom3 <- group_by(ecom2, device)
ecom3
```

## Summarise Data

The next step is to compute the total order value and total order items for each 
device. i.e. we need to reduce the order value and order items data to a single 
summary. We can achieve this using `summarise()`. As usual, the first argument 
is the name of a data set and the subsequent arguments are functions that can 
summarise data. For example, we can use `min`, `max`, `sum`, `mean` etc.

Let us compute the average number of pages browsed by referrer type:

- split data by `referrer` type
- compute the average number of pages using `mean`

```{r sumeg}
# split data by referrer type
step_1 <- group_by(ecom, referrer)

# compute average number of pages
step_2 <- summarise(step_1, mean(n_pages))
step_2
```

Now let us compute both the `mean` and the `median`.

```{r sumeg2}
# split data by referrer type
step_1 <- group_by(ecom, referrer)

# compute average number of pages
step_2 <- summarise(step_1, mean(n_pages), median(n_pages))
step_2
```

Another way to achieve the above result is to use the 
`summarise_all()` function. How does that work? It generates the specified 
summary for all the columns in the data set except for the column based on 
which the data has been grouped or split. So we need to ensure that the data 
does not have any irrelevant columns.

- split data by `referrer` type
- select `order_value` and `order_items`
- compute the average number of pages by applying the `mean` function to all 
the columns


```{r sumall_eg}
# select relevant columns
step_1 <- select(ecom, referrer, order_value, order_items)

# split data by referrer type
step_2 <- group_by(step_1, referrer)

# compute average number of pages
step_3 <- summarise_all(step_2, funs(mean))
step_3
```

Let us compute `mean` and `median` number of pages for each referre type using 
`summarise_all`.

```{r sumall_eg2}
# select relevant columns
step_1 <- select(ecom, referrer, order_value, order_items)

# split data by referrer type
step_2 <- group_by(step_1, referrer)

# compute mean and median number of pages
step_3 <- summarise_all(step_2, funs(mean, median))
step_3
```


##### Case Study

So far, we have split the data based on the `device` type and we have selected 2 
columns,  `order_value` and `order_items`. We need the sum of order value and 
order items. What function can we use to obtain them? The `sum()` function will 
generate the sum of the values and hence we will use it inside the `summarise()` 
function. Remember, we need to provide a name to the summary being generated. 

```{r dplyr13}
ecom4 <- summarise(ecom3, total_value = sum(order_value),
          total_items = sum(order_items))
ecom4
```

There you go, we have the total order value and total order items for each 
device type. If we use `summarise_all()`, it will generate the summary for the 
selected columns based on the function specified. To specify the functions, we 
need to use another argument `funs` and it can take any number of valid 
functions.

```{r dplyr14}
ecom4 <- summarise_all(ecom3, funs(sum))
ecom4
```

## Create Columns

To create a new column, we will use `mutate()`. The first argument is the name 
of the data set and the subsequent arguments are expressions for creating new 
columns based out of existing columns. 

Let us add a new column `avg_page_time` i.e. time on site divided by number of
pages visited.

```{r mutateg}
# select duration and n_pages from ecom
mutate_1 <- select(ecom, n_pages, duration)
mutate(mutate_1, avg_page_time = duration / n_pages)
```

We can create new columns based on other columns created using `mutate`. Let us 
create another column `sqrt_avg_page_time` i.e. square root of the average time
on page using `avg_page_time`.

```{r mutateg2}
mutate(mutate_1,
       avg_page_time = duration / n_pages,
       sqrt_avg_page_time = sqrt(avg_page_time))
```

##### Case Study

Back to our case study, from the last step we have the total order value and 
total order items for each device category and can compute the AOV. We will 
create a new column to store AOV. 

```{r dplyr15}
ecom5 <- mutate(ecom4, aov = order_value / order_items)
ecom5
```

## Select Columns

The last step is to select the relevant columns. We will select the device type 
and the corresponding aov while getting rid of other columns. Use `select()`
to extract the relevant columns.

```{r dplyr16}
ecom6 <- select(ecom5, device, aov)
ecom6
```

## Arrange Data

Arranging data in ascending or descending order is one of the most common tasks
in data manipulation. We can use `arrange` to arrange data by different columns.
Let us say we want to arrange data by the number of pages browsed.

```{r arrange_eg}
arrange(ecom, n_pages)
```

If we want to arrange the data in descending order, we can use `desc()`. Let us 
arrange the data in descending order.

```{r arrange_eg2}
arrange(ecom , desc(n_pages))
```

Data can be arranged by multiple variables as well. Let us arrange data first by 
number of visits and then by number of pages in a descending order.

```{r arrange_eg3}
arrange(ecom, n_visit, desc(n_pages))
```

##### Case Study

If you observe `ecom6`, the `aov` column is arranged in descending order. 

```{r arrange4}
arrange(ecom6, aov)
```

## AOV by Devices

Let us combine all the code from the above steps:

```{r dplyr17}
ecom1 <- filter(ecom, purchase == "true")
ecom2 <- select(ecom1, device, order_value, order_items)
ecom3 <- group_by(ecom2, device)
ecom4 <- summarise_all(ecom3, funs(sum))
ecom5 <- mutate(ecom4, aov = order_value / order_items)
ecom6 <- select(ecom5, device, aov)
ecom7 <- arrange(ecom6, aov)
ecom7
```

If you observe, at each step we create a new variable(data frame) and then use 
it as an input in the next step i.e. the output from one step becomes the input 
for the next. Can we achieve the final outcome i.e. `ecom7` without creating the 
intermediate data (ecom1 - ecom6)? Yes, we can. We will use the `%>%` operator 
to chain the steps and get rid of the intermediate data. 

```{r dplyr18}
ecom %>%
  filter(purchase == 'true') %>%
  select(device, order_value, order_items) %>%
  group_by(device) %>%
  summarise_all(funs(sum)) %>%
  mutate(
    aov = order_value / order_items
  ) %>%
  select(device, aov) %>%
  arrange(aov)
```

Below we map the description of each step to dplyr verbs.

<center>
<img src = "/post/aov_flow_map_canva.png" width = "1200" height = "800">
</center>

## Your Turn

- what is the average number of pages visited by purchasers and non-purchasers?
- what is the average time on site for purchasers vs non-purchasers?
- what is the average number of pages visited by purchasers and non-purchasers using mobile?
