---
title: 'Legend - Part 2'
date: '2017-07-06'
slug: legend-part-2
topics:
  - data-visualization
tags:
  - ggplot2
draft: yes
---

### Introduction

<hr>

This is the 14th post in the series **Elegant Data Visualization with ggplot2**. In the previos [post](https://rsquaredacademy.github.io/blog/post/guides-legends/), we learnt how to modify the legend of plots when aesthetics are mapped to variables. In this post, we will continue to explore different ways to modify/customize the legend of plots.


```{r gui1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

If the aesthetics are mapped to variables, ggplot2 automatically creates legends wherever applicable. 
You may want to modify the appearance of legends. In this module, we will learn to modify the legends when
`fill` is mapped to categorical variables.

Modify legend when `fill` is mapped to variables. Specifically

- title
- breaks
- limits
- labels
- values

<br>

### Libraries, Code & Data

<hr>

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/ef2b665173cf7cfe17093f3dcb052035).

<br>

### Plot

<hr>

```{r leg22}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22)
```

<br>

### Title

<hr>

```{r leg23}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(name = "Cylinders", 
    values = c("red", "blue", "green"))
```

<br>

### Values

<hr>

```{r leg24}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"))
```

<br>

### Labels

<hr>

```{r leg25}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'))
```

<br>

### Limits

<hr>

```{r leg26}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    limits = c(4, 6, 8))
```

<br>

### Breaks

<hr>

```{r leg27}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    breaks = c(4, 6, 8))
```

<br>

### Putting it all together...

<hr>

```{r leg28}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(name = "Cylinders", values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'), limits = c(4, 6, 8), breaks = c(4, 6, 8))
```

<br>

### Summary

<hr>

<br>

### Up Next..

<hr>

In the next [post](https://rsquaredacademy.github.io/blog/post/legend-part-3/), we will learn to modify the legend when 
`shape` is mapped to categorical variables.

<br>


