---
title: Importing Data into R - Part 2
author: Aravind Hebbali
date: '2017-11-12'
slug: importing-data-into-r-part-2
categories:
  - data wrangling
tags:
  - import
  - readxl
---

## Introduction

This is the second post in the series **Importing Data into R**. In the previous [post](), we explored reading data from flat/delimited files. In this post, we will:

- list sheets in an excel file
- read data from an excel sheet
- read specific cells from an excel sheet
- read specific rows
- read specific columns
- read data from
		- SAS 
		- SPSS
		- STATA

## Libraries, Data & Code

We will use the [readxl](http://readr.tidyverse.org/index.html) package. It has no external dependencies as compared to other packages available for reading data from Excel. The data sets can be downloaded from [here](https://github.com/rsquaredacademy/datasets) and the codes from [here](https://gist.github.com/aravindhebbali/65dc2af57b8eda098f2b878095c5aaa4).

```{r library, message=FALSE}
library(readxl)
```

## List Sheets

Before we read data from an excel file, let us see how many sheets are present using `excel_sheets()`.

```{r list_sheets}
excel_sheets('sample.xls')
```

## Read Sheet

To read data from a particular sheet, use `read_excel()` and specify the file name and the sheet number. Below is a simple example:

```{r import56, echo=TRUE, results='markup', message=FALSE}
read_excel('sample.xls', sheet = 1)
```

## Read Specific Cells

To read data from specific cells or a range of cells, use the `range` arguments and specify the range of cells from which we want to read data. For example, to read data from first 4 rows of columns **B** and **C**, we will specify the range as `"B1:C4"`.

```{r import58, echo=TRUE, results='markup', message=FALSE}
read_excel('sample.xls', sheet = 1, range = "B1:C4")
```

To read data from first 5 rows of columns **A**, **B** and **C**, we will specify the range as `"A1:C5"`.

```{r import59, echo=TRUE, results='markup', message=FALSE}
readxl::read_excel('sample.xls', sheet = 1, range = "A1:C5")
```

Another way to read specific cells is by providing a particular cell and then specifying the number of rows and columns keeping that cell as anchorage. In the below example, we want to read 3 rows and 2 columns starting from the cell `A4`.

```{r import61, echo=TRUE, results='markup', message=FALSE}
readxl::read_excel('sample.xls', sheet = 1, col_names = FALSE,
  range = anchored("A4", dim = c(3, 2)))
```

Use cell_limits to specify one end of the rectangle such as top left or top right.


```{r import65, echo=TRUE, results='markup', message=FALSE}
readxl::read_excel('sample.xls', sheet = 1,
  range = cell_limits(c(1, 2), c(NA, NA)))
```

```{r import66, echo=TRUE, results='markup', message=FALSE}
readxl::read_excel('sample.xls', sheet = 1,
  range = cell_limits(c(1, NA), c(NA, 2)))
```

## Specify Rows

Use `cell_rows()` to specify the rows from which data must be read. In the below example, we want to read the first 4 rows of data from the file.

```{r import62, echo=TRUE, results='markup', message=FALSE}
readxl::read_excel('sample.xls', sheet = 1, range = cell_rows(1:4))
```

## Specify Columns

Use `cell_cols()` to specify the columns from which data must be read. In the below example, we want to read the 2nd and 3rd column from the file.

```{r import63, echo=TRUE, results='markup', message=FALSE}
readxl::read_excel('sample.xls', sheet = 1, range = cell_cols(2:3))
```

## Statistical Softwares

We will use the [haven]() package to read data from files of other statistical softwares such as:

- SAS
- SPSS
- STATA

### Library

```{r haven, message=FALSE}
library(haven)
```

### STATA

```{r import68, echo=TRUE, results='markup', message=FALSE}
read_stata('airline.dta')  
```

### SPSS

```{r import71, echo=TRUE, results='markup', message=FALSE}
read_spss('employee.sav')  
```

### SAS

```{r import73, echo=TRUE, results='markup', message=FALSE}
read_sas('airline.sas7bdat')
```