---
title: Data Wrangling with dplyr - Part 3
author: Aravind Hebbali
date: '2017-12-24'
slug: data-wrangling-with-dplyr-part-3
categories:
  - data wrangling
tags:
  - dplyr
---

## Introduction

In the previous [post](https://blog.aravindhebbali.com/2017/12/25/data-wrangling-with-dplyr-part-2/), 
we learnt to combine tables using dplyr. In this post, we will explore a set of 
helper functions in order to:

- extract unique rows 
- rename columns
- sample data
- extract columns
- slice rows
- arrange rows
- compare tables
- extract/mutate data using predicate functions
- count observations for different levels of a variable

## Libraries, Code & Data

We will use the following packages:

- [dplyr](http://dplyr.tidyverse.org/index.html) 
- [readr](http://readr.tidyverse.org/index.html)  

The data sets can be downloaded from [here](https://github.com/rsquaredacademy/datasets) 
and the codes from [here](https://gist.github.com/aravindhebbali/55c4f40476028c09949b73af97bb1619).

```{r dplyr1, message=FALSE}
library(dplyr)
library(readr)
```

## Case Study

Let us look at a case study (e-commerce data) and see how we can use dplyr 
helper functions to answer questions we have about and to modify/transform the 
underlying data set. 

### Data

```{r show, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
ecom
```

### Data Dictionary

- id: row id
- referrer: referrer website/search engine
- os: operating system
- browser: browser
- device: device used to visit the website
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- repeat: frequency of visits
- country: country of origin
- purchase: whether visitor purchased
- order_value: order value of visitor (in dollars)

## Data Sanitization

Let us ensure that the data is sanitized by checking the sources of traffic
and devices used to visit the site. We will use `distinct` to examine the
values in the `referrer` column

```{r dplyr2}
ecom %>%
  distinct(referrer)
```

and the `device` column as well.

```{r dplyr30}
ecom %>%
  distinct(device)
```

## Data Tabulation

Let us now look at the proportion or share of visits driven by different sources
of traffic. 

```{r dplyr14}
ecom %>%
  group_by(referrer) %>%
  tally()
```

We would also like to know the number of bouncers driven by the different 
sources of traffic.

```{r dplyr15}
ecom %>%
  group_by(referrer, bouncers) %>%
  tally()
```

Let us look at how many conversions happen across different devices.

```{r dplyr25}
ecom %>%
  group_by(device, purchase) %>%
  tally() %>%
  filter(purchase == 'true')
```

Another way to extract the above information is by using `count`

```{r dplyr17}
ecom %>%
  count(referrer, purchase) %>%
  filter(purchase == "true")
```

## Sampling Data

dplyr offers sampling functions which allow us to specify either the
number or percentage of observations. `sample_n()` allows sampling a specific
number of observations.  

```{r dplyr4}
ecom %>%
  sample_n(700)
```

We can combine the sampling functions with other dplyr functions as shown below 
where we sample observation after grouping them according to the source of 
traffic.

```{r dplyr5}
ecom %>%
  group_by(referrer) %>%
  sample_n(100)
```

`sample_frac()` allows a specific percentage of observations.

```{r dplyr6}
ecom %>%
  sample_frac(size = 0.7)
```

## Data Extraction

In the first [post](https://blog.aravindhebbali.com/2017/12/25/data-wrangling-with-dplyr-part-1/),
we had observed that dplyr verbs always returned a tibble. What if you want to
extract a specific column or a bunch of rows but not as a tibble? 

Use `pull` to extract columns either by name or position. It will return a 
vector. In the below example, we extract the `device` column as a vector. I am 
using `head` in addition to limit the output printed.

```{r dplyr8}
ecom %>%
  pull(device) %>%
  head
```

Let us extract the first column from `ecom` using column position instead of 
name.

```{r dplyr9}
ecom %>%
  pull(1) %>%
  head
```

You can use `-` before the column position to indicate the position in reverse.
The below example extracts data from the last column.

```{r dplyr10}
ecom %>%
  pull(-1) %>%
  head
```

Let us now look at extracting rows using `slice()`. In the below example, we 
extract data starting from the 5th row and upto the 15th row.

```{r dplyr11}
ecom %>%
  slice(5:15)
```

Use `n()` inside `slice()` to extract the last row.

```{r dplyr12}
ecom %>%
  slice(n())
```

## Between

`between()` allows us to test if the values in a column lie between two specific
values. In the below example, we check how many visits browsed pages between 5 
and 15.

```{r dplyr19}
ecom_sample <-
  ecom %>%
  sample_n(30)
  
ecom_sample %>%
  pull(n_pages) %>%
  between(5, 15) 
```

## Case When

`case_when()` is an alternative to `if else`. It allows us to lay down the 
conditions clearly and makes the code more readable. In the below example, we
create a new column `repeat_visit` from `n_visit` (the number of previous visits).

```{r dplyr20}
ecom %>%
  mutate(
    repeat_visit = case_when(
      n_visit > 0 ~ TRUE,
      TRUE ~ FALSE
    )
  ) %>%
  select(n_visit, repeat_visit) 
```

