---
title: 'ggplot2: Legend - Part 5'
author: Aravind Hebbali
twitterImg: /img/gg_leg_part_1.png
description: "ggplot2 - Legend Part 5"
date: '2018-04-01'
slug: legend-part-5
topics:
  - data-visualization
tags:
  - ggplot2
---

### Introduction

<hr>

This is the 17th post in the series **Elegant Data Visualization with ggplot2**. 
In the previous post, we learnt how to modify the legend of plot when `size` is 
mapped to continuous variable. In this post, we will learn to modify the 
following using `scale_alpha_continuous()` when `alpha` or transparency is 
mapped to variables:

- title
- breaks
- limits
- range
- labels
- values

```{r gui1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

<br>

### Libraries, Code & Data

<hr>

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/ae9f6afd3c4818b3778ea008da7d2e92).

<br>

### Plot

<hr>

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the transparency of 
the points to the `hp` variable. Remember, `alpha` must always be mapped to a 
continuous variable.


```{r leg2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue')
```

As you can see, the legend acts as a guide for the `alpha` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

<br>

### Title

<hr>

The title of the legend (`hp`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Horsepower` using the `name` argument.

```{r leg3, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous("Horsepower")
```

<br>

### Breaks

<hr>

When the range of the variable mapped to size is large, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
certain midpoints (125, 200, 275) of the mapped variable.

```{r leg4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(breaks = c(125, 200, 275))
```

<br>

### Limits

<hr>

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars whose horsepower is between `100` and `350`. 
One way to approach this would be to filter the data using `filter` from dplyr 
and then visualize it. Instead, we will use the `limits` argument and filter 
the data for visualization.


```{r leg5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(limits = c(100, 350))
```

<br>

### Range

<hr>

The range of the transparency of points can be modified using the `range` 
argument. We need to specify a lower and upper range using a numeric vector. 
In the below example, we use `range` and supply the lower and upper limits as 
`0.4` and `0.8`. The transparency of the points will now lie between `0.4` and 
`0.8` only.

```{r leg6, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(range = c(0.4, 0.8))
```

<br>

### Labels

<hr>

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to "1 Hundred", "2 Hundred" and "3 Hundred" in the next example. 
Ensure that the labels are intuitive and easy to interpret for the end user of 
the plot.

```{r leg7, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", 
      "3 Hundred"))
```

<br>

### Putting it all together

<hr>

```{r leg8, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous("Horsepower", breaks = c(100, 200, 300),
    limits = c(100, 350), range = c(0.4, 0.8),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```

<br>

### Summary

<hr>

In this post, we learnt to modify the following aspects of legends: 

- title
- breaks
- range
- limits
- labels
- values

<br>

### Up Next..

<hr>

In the next post, we will learn how to modify the title, label and bar of the legend.

<br>