---
title: Working with Categorical Data using forcats
author: Aravind Hebbali
date: '2018-01-21'
slug: working-with-categorical-data-using-forcats
categories:
  - data wrangling
tags:
  - forcats
  - factor
  - categorical
---

## Introduction

In this post, we will learn to work with categorical/qualitative data in R using [forcats](http://forcats.tidyverse.org). Let us begin by installing and loading forcats and a set of other pacakges we will be using.

## Libraries &  Code

We will use the following packages: 

- [forcats](http://forcats.tidyverse.org/index.html) 
- [dplyr](http://dplyr.tidyverse.org/index.html) 
- [magrittr](http://magrittr.tidyverse.org/index.html) 
- [ggplot2](http://ggplot2.tidyverse.org/index.html) 
- [tibbe](http://tibble.tidyverse.org/index.html) 
- [purrr](http://purrr.tidyverse.org/index.html) 
- and [readr](http://readr.tidyverse.org/index.html)  

The codes from [here](https://gist.github.com/aravindhebbali/85fac536f563ae3fd8e2605fd56a7086).

```{r cat1, message=FALSE}
library(forcats)
library(tibble)
library(magrittr)
library(purrr)
library(dplyr)
library(ggplot2)
library(readr)
```

## Case Study

We will use a case study to explore the various features of the forcats package. You can download the data for the case study from [here](https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv) or directly import the data using the readr package. We will do the following in this case study:

- compute the frequency of different referrers
- plot average number of pages browsed for different referrers
- collapse referrers with low sample size into a single group
- club traffic from social media websites into a new category
- group referrers with traffic below a threshold into a single category
 
### Data

```{r show, message=FALSE}
ecom <- read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv',
                 col_types = list(col_integer(), 
                                  col_factor(levels = c("bing", "direct", "social", "yahoo", "google")), 
                                  col_factor(levels = c("laptop", "mobile", "tablet")), 
                                  col_logical(), col_number(),
                                  col_number(), col_number(), col_character(), 
                                  col_logical(), col_number(), col_number()))

ecom
```

### Data Dictionary

- id: row id
- referrer: referrer website/search engine
- os: operating system
- browser: browser
- device: device used to visit the website
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- repeat: frequency of visits
- country: country of origin
- purchase: whether visitor purchased
- order_value: order value of visitor (in dollars)

## Tabulate Referrers

Let us look at the traffic driven by different referrers. 

```{r cat2}
ecom %>%
  pull(referrer) %>%
  fct_count
```

If you want to sort the output in descending order, use `sort` and set it to `TRUE`.

```{r cat2a}
ecom %>%
  pull(referrer) %>%
  fct_count(sort = TRUE)
```

## Reorder Referrers

We want to examine the average number of pages visited by each referrer type. 

```{r cat3}
refer_summary <- 
  ecom %>%
  group_by(referrer) %>%
  summarise(
    page = mean(n_pages),
    tos = mean(duration),
    n = n()
  )

refer_summary
```

Let us plot the average number of pages visited by each referrer type. 

```{r cat10, fig.align='center', fig.width=6, fig.height=4}
refer_summary %>%
  ggplot() +
  geom_point(aes(page, referrer))
```

Use `fct_reorder` to reorder the referrer types by the average number of pages
visited.

```{r cat11, fig.align='center', fig.width=6, fig.height=4}
refer_summary %>%
  ggplot() +
  geom_point(aes(page, fct_reorder(referrer, page)))
```

## Plot Referrer Frequency (Descending Order)

Let us look at the distribution of the referreres.

```{r cat21}
ecom %>%
  pull(referrer) %>%
  fct_count(sort = TRUE)
```

Use `fct_unique` to view the categories or levels of the referrer variable.

```{r cat22}
ecom %>%
  pull(referrer) %>%
  fct_unique
```

Since we want to plot the referrers in descending order of frequency, we will
use `fct_infreq()` to reorder by frequency.

```{r cat23}
ecom %>%
  pull(referrer) %>%
  fct_infreq %>%
  fct_unique
```

Now that we have reordered the referrers by frequency, let us plot them.

```{r cat4, fig.align='center', fig.width=6, fig.height=4}
ecom %>%
  mutate(
    ref = referrer %>% 
      fct_infreq
  ) %>%
  ggplot(aes(ref)) +
  geom_bar()
```

## Plot Referrer Frequency (Ascending Order)

Let us look at the categories of the referrer variable. 

```{r cat24}
ecom %>%
  pull(referrer) %>%
  fct_unique
```

Since we want to plot the referrers in ascending order of frequency, we will 
use `fct_rev()` to reverse the order.

```{r cat25}
ecom %>%
  pull(referrer) %>%
  fct_rev %>%
  fct_unique
```

Let us reorder the referrers by frequency first and then reverse the order 
before plotting their frequencies.

```{r cat5, fig.align='center', fig.width=6, fig.height=4}
ecom %>%
  mutate(
    ref = referrer %>% 
      fct_infreq %>% 
      fct_rev
  ) %>%
  ggplot(aes(ref)) +
  geom_bar()
```

## Case Study 2

### Data 

```{r import2, message=FALSE}
traffic <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web_traffic.csv',
  col_types = list(col_factor(levels = c("affiliates", "bing", "direct", "facebook", "yahoo", "google",
    "instagram", "twitter", "unknown"))))

traffic
```

## Tabulate Referrer

Let us compute the traffic driven by different referrers using `fct_count`.

```{r cat6}
traffic %>%
  pull(traffics) %>% 
  fct_count
```

## Collapse Referrer Categories

We want to group the referrers into 2 categories:

- social
- search

Use `fct_collapse()` to group categories.

```{r cat7}
traffic %>%
  pull(traffics) %>%
  fct_collapse(
    social = c("facebook", "twitter", "instagram"),
    search = c("google", "bing", "yahoo")
  ) %>% 
  fct_count
```

## Lump Infrequent Referrer Types

Let us group together referrer types that drive low traffic to the website.
Use `fct_lump()` to lump together categories.

```{r cat8, collapse = TRUE}
traffic %>%
  pull(traffics) %>% 
  fct_count

traffic %>%
  pull(traffics) %>% 
  fct_lump %>% 
  table
```

## Retain top 3 referrers

We want to retain the top 3 referrers and combine the rest of them into a
single category.

```{r cat9, echo=FALSE}
traffic %>%
  pull(traffics) %>% 
  fct_count %>% 
  arrange(desc(n))
```

Use `fct_lump()` and set the argument `n` to `3` indicating we want to
retain top 3 categories and combine the rest.

```{r cat17}
traffic %>% 
  pull(traffics) %>% 
  fct_lump(n = 3) %>% 
  table
```

## Lump Referrer Types with less than 10% traffic

Let us combine referrers that drive less than 10% traffic to the website.

```{r cat12, echo=FALSE}
traffic %>% 
  pull(traffics) %>% 
  fct_count %>%
  mutate(
    percent = round((n / sum(n)) * 100, 2)
  )
```

Since we are looking at proportion of traffic driven to the website and
not the actual numbers, we use the `prop` argument and set it to `0.1`,
indicating that we want to retain only those categories which have a
proportion of more than 10% and combine the rest.

```{r cat16}
traffic %>%
  pull(traffics) %>% 
  fct_lump(prop = 0.1) %>% 
  table
```

## Retain 3 Referrer Types with lowest traffic

What if we want to retain 3 referrers which drive the lowest traffic 
to the website and combine the rest? 

```{r cat14, echo=FALSE}
traffic %>%
  pull(traffics) %>% 
  fct_count %>% 
  arrange(n)
```

We will still use the `n` argument but instead of specifying `3`, we now specify `-3`. 

```{r cat19}
traffic %>% 
  pull(traffics) %>% 
  fct_lump(n = -3) %>% 
  table
```

## Retain 3 Referrer Types with less than 10% traffic

Let us see how to retain referrers that drive less than 10 % traffic to 
the website and combine the rest into a single group. 

```{r cat15, echo=FALSE}
traffic %>%
  pull(traffics) %>% 
  fct_count %>%
  mutate(
    percent = round((n / sum(n)) * 100, 2)
  )
```

Instead of setting `prop` to `0.1`, we will set it to `-0.1`.

```{r cat20}
traffic %>% 
  pull(traffics) %>% 
  fct_lump(prop = -0.1) %>% 
  table
```