---
title: SQL for Data Science - Part 2
date: '2018-02-18'
slug: sql-for-data-science-part-2
categories:
  - database
tags:
  - sql
  - sqlite
---

```{r setup, eval=TRUE, message=FALSE, echo=FALSE}
library(dplyr)
library(DBI)
library(RSQLite)
knitr::opts_chunk$set(R.options=list(max.print=50))
```

#### Introduction

This is the fourth post in the series **R & Databases**. You can find the links to the other two posts of this series below:

- [Quick Guide: R & SQLite](https://blog.aravindhebbali.com/2018/01/28/quick-guide-r-sqlite/)
- [Data Wrangling with dbplyr](https://blog.aravindhebbali.com/2018/02/04/data-wrangling-with-dbplyr/)
- [SQL for Data Science - Part 1](https://blog.aravindhebbali.com/2018/02/11/sql-for-data-science-part-1/)

In this post, we will learn to

- aggregate data
- order data
- and group data

### Libraries, Code & Data

We will use the following libraries in this post:

- [DBI](http://readr.tidyverse.org/)
- [RSQLite](https://rstats-db.github.io/RSQLite/)
- [dbplyr](http://dbplyr.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/691dba8e13873c648bfdb91aba65d4d7).

#### Set Up

```{r setup2, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con, ecom)
```

#### Aggregate 

Let us combine the aggregate statements with WHERE statement to filter data.

- SUM()  : returns the total sum of a numeric column

```{r sql1}
dbGetQuery(con, "SELECT SUM(n_visit) FROM ecom")

dbGetQuery(con, "SELECT SUM(n_visit) 
	             FROM ecom 
	             WHERE n_visit > 5")
```

- AVG()  : returns the average value of a numeric column

```{r sql2}
dbGetQuery(con, "SELECT AVG(n_visit) FROM ecom")

dbGetQuery(con, "SELECT AVG(n_visit) 
	             FROM ecom 
	             WHERE country LIKE 'P%'")
```

- MAX()  : returns the largest value of the selected column

```{r sql3}
dbGetQuery(con, "SELECT MAX(n_visit) FROM ecom")

dbGetQuery(con, "SELECT MAX(n_visit) 
	             FROM ecom 
	             WHERE device == 'tablet'")
```

- MIN()  : returns the smallest value of the selected column

```{r sql4}
dbGetQuery(con, "SELECT MIN(n_visit) FROM ecom")

dbGetQuery(con, "SELECT MIN(n_visit) 
	             FROM ecom 
	             WHERE duration BETWEEN 600 AND 900")
```

#### Alias 

SQL aliases are used to give a table, or a column in a table, a temporary name. They are often used to make column names more readable. An alias only exists for the duration of the query. Below are a
few examples:

```{r sql9}
dbGetQuery(con, "SELECT AVG(n_visit) AS avg_mobile 
	             FROM ecom 
	             WHERE device == 'mobile'")

dbGetQuery(con, "SELECT MAX(n_visit) AS max_visit 
	             FROM ecom")

dbGetQuery(con, "SELECT MIN(duration) AS min_duration 
	             FROM ecom")
```

#### Order By

The ORDER BY keyword is used to sort the records in ascending or descending order. By default, the records are sorted in ascending order. Use the DESC keyword if you want to sort the records in descending order, 

```{r sql12}
dbGetQuery(con, "SELECT * 
	             FROM ecom 
	             ORDER BY country")

dbGetQuery(con, "SELECT *  
	             FROM ecom 
	             ORDER BY duration")

dbGetQuery(con, "SELECT * 
	             FROM ecom 
	             ORDER BY n_visit DESC")
```

#### Group By 

The GROUP BY statement is used with aggregate functions (COUNT, MAX, MIN, SUM, AVG) to group the result by one or more columns.

```{r sql15}
dbGetQuery(con, "SELECT device, 
	             count(*) AS visits 
	             FROM ecom 
	             GROUP BY device 
	             ORDER by visits DESC")

dbGetQuery(con, "SELECT device, MAX(duration) AS max_duration 
	             FROM ecom 
	             GROUP BY device 
	             ORDER by max_duration DESC")
```

### Summary

In this post, we learnt to 

- aggregate data 
- order data
- and group data


