---
title: 'ggplot2: Legend - Part 2'
author: Aravind Hebbali
twitterImg: /img/gg_leg_part_2.png
description: "ggplot2 - Legend Part 2"
date: '2018-02-24'
slug: legend-part-2
topics:
  - data-visualization
tags:
  - ggplot2
---

### Introduction

<hr>

This is the 14th post in the series **Elegant Data Visualization with ggplot2**. 
In the previous post, we learnt how to modify the legend of plots when 
aesthetics are mapped to variables. In this post, we will continue to explore 
different ways to modify/customize the legend of plots.


```{r gui1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

Specifically, we will learn to modify the following using `scale_fill_manual()` 
when `fill` is mapped to categorical variables:

- title
- breaks
- limits
- labels
- values

<br>

### Libraries, Code & Data

<hr>

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/ef2b665173cf7cfe17093f3dcb052035).

<br>

### Plot

<hr>

Let us start with a scatter plot examining the relationship between 
displacement and miles per gallon from the mtcars data set. We will map `fill`
to the `cyl` variable. 

```{r leg22}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22)
```

As you can see, the legend acts as a guide for the `color` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

<br>

### Title

<hr>

The title of the legend (`factor(cyl)`) is not very intuitive. If the user 
does not know the underlying data, they will not be able to make any sense out 
of it. Let us change it to `Cylinders` using the `name` argument.


```{r leg23}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(name = "Cylinders", 
    values = c("red", "blue", "green"))
```

<br>

### Values

<hr>

To change the default colors in the legend, use the `values` argument and
supply a character vector of color names. The number of colors specified 
must be equal to the number of levels in the categorical variable mapped.
In the below example, `cyl` has 3 levels (4, 6, 8) and hence we have specified
3 colors.

```{r leg24}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"))
```

<br>

### Labels

<hr>

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to `Four`, `Six` and `Eight` in the next example. Ensure that
the labels are intuitive and easy to interpret for the end user of the plot.

```{r leg25}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'))
```

<br>

### Limits

<hr>

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars with at least 6 cylinders. One way to approach this
would be to filter the data using `filter` from dplyr and then visualize it. 
Instead, we will use the `limits` argument and filter the data for visualization.

```{r leg26}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    limits = c(6, 8))
```

As you can see above, `ggplot2` returns a warning message indicating data related
to 4 cylinders has been dropped. If you observe the legend, it now represents 
only 4 and 6 cylinders.

<br>

### Breaks

<hr>

When there are large number of levels in the mapped variable, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
two levels (4, 8) of the mapped variable.

```{r leg27}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    breaks = c(4, 8))
```

<br>

### Putting it all together...

<hr>

```{r leg28}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(name = "Cylinders", values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'), limits = c(4, 6, 8), breaks = c(4, 6, 8))
```

<br>

### Summary

<hr>

In this post, we learnt to modify the following aspects of legends: 

- title
- breaks
- limits
- labels
- values

<br>

### Up Next..

<hr>

In the next post, we will learn to modify the legend when `shape` is mapped to categorical variables.

<br>


