---
title: 'ggplot2: Legend - Part 6'
author: Aravind Hebbali
twitterImg: /img/gg_leg_part_1.png
description: "ggplot2 - Legend Part 6"
date: '2018-04-13'
slug: legend-part-6
topics:
  - data-visualization
tags:
  - ggplot2
---


### Introduction

<hr>

This is the 18th post in the series **Elegant Data Visualization with ggplot2**. 
In the previous post, we learnt how to modify the legend of plot when `alpha` 
is mapped to a categorical variable. In this post, we will learn to modify legend

- title
- label
- and bar

So far, we have learnt to modify the components of a legend using `scale_*` 
family of functions. Now, we will use the `guide` argument and supply it 
values using the `guide_legend()` function.

<br>

```{r load, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

<br>

### Libraries, Code & Data

<hr>

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/a77a836e7e06c61c28ba7db3a69e1510).

<br>

### Title

#### Title Alignment

<hr>

The horizontal alignment of the title can be managed using the `title.hjust` 
argument. It can take any value between `0` and `1`.

- 0 (left)
- 1 (right)

In the below example, we align the title to the center by assigning the value
`0.5`.

```{r leg2, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5))
```

<br>

#### Title Alignment (Vertical)

<hr>

To manage the vertical alignment of the title, use `title.vjust`.

```{r leg22, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    title = "Horsepower", title.position = "top", title.vjust = 1))
```

<br>


#### Title Position

<hr>

The position of the title can be managed using `title.posiiton` argument. It 
can be positioned at:

- top
- bottom
- left
- right


```{r leg3, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top"))
```

<br>

## Label


### Label Position

<hr>

The position of the label can be managed using the `label.position` argument.
It can be positioned at:

- top
- bottom
- left
- right

In the below example, we position the label at right.

```{r leg4, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.position = "right"))
```

<br>

### Label Alignment

<hr>

The horizontal alignment of the label can be managed using the `label.hjust` 
argument. It can take any value between `0` and `1`.

- 0 (left)
- 1 (right)

In the below example, we align the label to the center by assigning the value
`0.5`.

- alignment 
    - 0 (left)
    - 1 (right)

```{r leg5, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.hjust = 0.5))
```

<br>

#### Labels Alignment (Vertical)

<hr>

The vertical alignment of the label can be managed using the `label.vjust` 
argument.

```{r leg23, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    label.vjust = 0.8))
```

<br>

### Direction

<hr>

The direction of the label can be either horizontal or veritcal and it can be 
set using the `direction` argument. 

```{r leg6, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(direction = "horizontal"))
```

<br>

### Rows

<hr>

The label can be spread across multiple rows using the `nrow` argument. In the
below example, the label is spread across 2 rows.

```{r leg7, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(nrow = 2))
```

<br>

### Reverse

<hr>

The order of the labels can be reversed using the `reverse` argument. We need
to supply logical values i.e. either `TRUE` or `FALSE`. If `TRUE`, the order
will be reversed.

```{r leg8, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(reverse = TRUE))
```

<br>

### Putting it all together...

<hr>

```{r leg20, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top", label.position = "right", 
      direction = "horizontal", label.hjust = 0.5, nrow = 2, reverse = TRUE)
  )
```

<br>


### Legend Bar 

<hr>

So far we have looked at modifying components of the legend when it acts as a
guide for `color`, `fill` or `shape` i.e. when the aesthetics have been mapped 
to a categorical variable. In this section, you will learn about
`guide_colorbar()` which will allow us to modify the legend when the aesthetics
are mapped to a continuous variable.

### Plot

<hr>

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the color of the points
to the `hp` variable. 

```{r leg36}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp))
```

<br>

### Width

<hr>

The width of the bar can be modified using the `barwidth` argument. It is used
inside the `guide_colorbar()` function which itself is supplied to the `guide`
argument of `scale_color_continuous()`. 

```{r leg24, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barwidth = 10))
```

<br>

### Height

<hr>

Similarly, the height of the bar can be modified using the `barheight` argument.

```{r leg25, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barheight = 3))
```

<br>

### Bins

<hr>

The `nbin` argument allows us to specify the number of bins in the bar.

```{r leg26, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    nbin = 4))
```

<br>

### Ticks

<hr>

The ticks of the bar can be removed using the `ticks` argument and setting it 
to `FALSE`.

```{r leg28, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    ticks = FALSE))
```

<br>

### Upper/Lower Limits

<hr>

The upper and lower limits of the bars can be drawn or undrawn using the 
`draw.ulim` and `draw.llim` arguments. They both accept logical values.

```{r leg29, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    draw.ulim = TRUE, draw.llim = FALSE))
```

<br>

### Guides: Color, Shape & Size

<hr>

The `guides()` function can be used to create multiple legends to act as a
guide for `color`, `shape`, `size` etc. as shown below. First, we map color, 
shape and size to different variables. Next, in the `guides()` function, we
supply values to each of the above aesthetics to indicate the type of legend.

```{r leg34, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, 
    size = qsec, shape = factor(gear))) + 
  guides(color = "colorbar", shape = "legend", size = "legend")
```

<br>

### Guides: Title

<hr>

To modify the components of the different legends, we must use the 
`guide_*` family of functions. In the below example, we use `guide_colorbar()`
for the legend acting as guide for color mapped to a continuous variable and
`guide_legend()` for the legends acting as guide for shape/size mapped to 
categorical variables.

```{r leg35, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, size = wt, shape = factor(gear))) + 
  guides(color = guide_colorbar(title = "Horsepower"),
    shape = guide_legend(title = "Weight"), size = guide_legend(title = "Gear")
  )
```

<br>

### Summary

<hr>

In this post, we will learn to modify legend

- title
- label
- and bar

<br>

### Up Next..

<hr>

In the next post, we will learn faceting.

<br>