---
title: 'ggplot2: Histogram'
author: Aravind Hebbali
twitterImg: /img/gg_hist.png
description: "ggplot2 - Histograms"
date: '2018-01-19'
slug: ggplot2-histogram
topics:
  - ggplot2
tags:
  - ggplot2
  - visualization
---

### Introduction

This is the eleventh post in the series **Elegant Data Visualization with 
ggplot2**. In the previous post, we learnt to build box plots. In this post, 
we will learn to

- build histogram
- specify bins
- modify 
    - color
    - fill
    - alpha
    - bin width
    - line type
    - line size
- map aesthetics to variables

A histogram is a plot that can be used to examine the shape and spread of  continuous data. It looks very similar to a bar graph and can be used to detect outliers and skewness in data. The histogram graphically shows the following:

- center (location) of the data
- spread (dispersion) of the data
- skewness
- outliers
- presence of multiple modes

To construct a histogram, the data is split into intervals called bins. The intervals may or may not be equal sized. For each bin, the number of data points that fall into it are counted (frequency). The Y axis of the histogram represents the frequency and the X axis represents the variable. 

```{r hist, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

<br>

### Libraries, Code & Data

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) 
and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/674bc30cc1539d735bdc4e6210982d1d).

<br>

### Data 

```{r import, eval=TRUE, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
ecom
```

<br>

### Data Dictionary

- id: row id
- referrer: referrer website/search engine
- os: operating system
- browser: browser
- device: device used to visit the website
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- repeat: frequency of visits
- country: country of origin
- purchase: whether visitor purchased
- order_value: order value of visitor (in dollars)

<br>

#### Histogram

To create a histogram, we will use `geom_histogram()` and specify the variable
name within `aes()`. In the below example, we create histogram of the variable 
`n_visit`.

```{r hist2, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit))
```

<br>

#### Specify Bins

The default number of bins in ggplot2 is `30`. You can modify the number of 
bins using the `bins` argument. In the below example, we create a histogram
with 7 bins.

```{r hist4, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 7)
```

<br>

#### Aesthetics

Now that we know how to create a histogram, let us learn to modify its 
appearance. We will begin with the background color. Use the `fill` argument
to modify the background color of the histogram. In the below case, we change
the color of the histogram to 'blue'.

```{r hist3, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 7, fill = 'blue')
```

<br>

As we have learnt before, the transparency of the background color can be 
modified using the `alpha` argument. It can take any value between `0` and `1`.

```{r hist11, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 7, fill = 'blue', alpha = 0.3)
```

<br>

The color of the histogram border can be modified using the `color` argument. 
The color can be specified either using its name or the associated hex code.

```{r hist8, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 7, fill = 'white', color = 'blue')
```
<br>

#### Putting it all together...

Let us modify the bins, the background and border color of the histogram in 
the below example.

```{r hist9, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 7, fill = 'blue', color = 'white')
```

<br>

#### Bin Width

Another way to control the number of bins in a histogram is by using the 
`binwidth` argument. In this case, we specify the width of the bins instead
of the number of bins. As you can see, in the below example, we do not use 
the `bins` argument when using the `binwidth` argument. You can use either of
them but not both.

```{r hist5, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), binwidth = 2, fill = 'blue', color = 'black')
```

<br>

#### Line Type

The line type of the histogram border can be modified using the `linetype` 
argument. It can take any integer value between `0` and `6`.

```{r hist6, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 5, fill = 'white', 
    color = 'blue', linetype = 3)
```

<br>

#### Line Size

Use the `size` argument to modify the width of the border of the histogram bins. 
It can take any value greater than `0`.

```{r hist10, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit), bins = 5, fill = 'white', 
    color = 'blue', size = 1.25)
```

<br>

#### Map Variables

You can map the aesthetics to variables as well. In the below example, we map
`fill` to the device variable. You can try mapping color, linetype and size to 
variables as well.

```{r hist7, fig.align='center'}
ggplot(ecom) +
  geom_histogram(aes(n_visit, fill = device), bins = 7)
```

<br>

### Summary

In this post, we learnt to:

- build histogram
- specify bins
- modify 
    - color
    - fill
    - alpha
    - bin width
    - line type
    - line size
- map aesthetics to variables


<br>

### Up Next..

In the next post, we will learn to modify the axes of a plot.

<br>