---
title: 'Working with Dates in R '
author: Aravind Hebbali
date: '2018-01-14'
slug: working-with-dates-in-r
categories:
  - data wrangling
tags:
  - lubridate
  - date
  - time
---

## Introduction

 In this post, we will learn to work with date/time data in R using [lubridate]( http://lubridate.tidyverse.org/), an R package that makes it easy to work with dates and time. 
 Let us begin by installing and loading the pacakge.
 
## Libraries, Code & Data

We will use the following packages: 

- [lubridate](http://lubridate.tidyverse.org/index.html) 
- [dplyr](http://dplyr.tidyverse.org/index.html)
- [magrittr](http://magrittr.tidyverse.org/index.html) 
- [readr](http://readr.tidyverse.org/index.html) 

The data sets can be downloaded from [here](https://github.com/rsquaredacademy/datasets) and the codes from [here](https://gist.github.com/aravindhebbali/7758b86c2bc13ff1e5d88d9d1c204f8c).

```{r lub1, message=FALSE}
library(lubridate)
library(dplyr)
library(magrittr)
library(readr)
```

## Quic Intro

#### Origin

Let us look at the origin for the numbering system used for date and time 
calculations in R.

```{r origin}
origin
```

#### Current Date/Time

Next, let us check out the current date, time and whether it occurs in the am 
or pm. `now()` returns the date time as well as the time zone whereas `today()`
will return only the current date. `am()` and `pm()` return `TRUE` or `FALSE`.

```{r today}
now()
today()
am(now())  
pm(now())
```

## Case Study

### Data

```{r import, eval=FALSE}
transact <- read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/transact.csv')
```

```{r show, echo=FALSE, eval=TRUE, message=FALSE}
transact <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/transact.csv')
transact
```

### Data Dictionary

The data set has 3 columns. All the dates are in the format (yyyy-mm-dd).

- Invoice: invoice date 
- Due: due date
- Payment: payment date

We will use the functions in the lubridate package to answer a few 
questions we have about the transact data.

- extract date, month and year from Due
- compute the number of days to settle invoice
- compute days over due
- check if due year is a leap year
- check when due day in february is 29, whether it is a leap year
- how many invoices were settled within due date
- how many invoices are due in each quarter
- what is the average duration between invoice date and payment date

## Extract Date, Month & Year from Due Date

The first thing we will learn is to extract the date, month and year. 

```{r lub18}
this_day <- as_date('2017-03-23')
day(this_day)
month(this_day)
year(this_day)
```

Let us now extract the date, month and year from the `Due` column.

```{r lub2}
transact %>%
  mutate(
    due_day = day(Due),
    due_month = month(Due),
    due_year = year(Due)
  )
```

## Compute days to settle invoice

Time to do some arithmetic with the dates. Let us calculate the duration of a 
course by subtracting the course start date from the course end date.

```{r lub19, collapse = TRUE}
course_start <- as_date('2017-04-12')

course_end <- as_date('2017-04-21')

course_duration <- course_end - course_start

course_duration
```

Let us estimate the number of days to settle the invoice by subtracting the
date of invoice from the date of payment.

```{r lub3}
transact %>%
  mutate(
    days_to_pay = Payment - Invoice
  )
```

## Compute days over due

How many of the invoices were settled post the due date? We can find this by:

- subtracting the due date from the payment date
- counting the number of rows where delay < 0 

```{r lub4}
transact %>%
  mutate(
    delay = Due - Payment
  ) %>%
  filter(delay < 0) %>%
  tally
```

## Is due year a leap year?

Just for fun, let us check if the due year happens to be a leap year.

```{r lub5}
transact %>%
  mutate(
    is_leap = leap_year(Due)
  )
```

## If due day is February 29, is it a leap year?

Let us do some data sanitization. If the due day happens to be February 29, 
let us ensure that the due year is a leap year. We will extract the following
from the due date:

- day
- month
- year

We will then create a new column `is_leap` which will have be set to `TRUE` if 
the year is a leap year else it will be set to `FALSE` and filter all the rows
where the month is 2 and the day is 29. Since we do not need 
the invoice or payment date, they will not be selected.

```{r lub6}
transact %>%
  mutate(
    due_day = day(Due),
    due_month = month(Due),
    due_year = year(Due),
    is_leap = leap_year(due_year)
  ) %>%
  select(-(Invoice), -(Payment)) %>%
  filter(due_month == 2 & due_day == 29)
```

## Shift Date

Time to shift some dates. We can shift a date by days, weeks or months. Let us 
shift the course start date by:

- 2 days
- 3 weeks
- 1 year

```{r lab40, collapse = TRUE}
course_start + days(2)

course_start + weeks(3)

course_start + years(1)
```

## Interval

Let us calculate the duration of the course using `interval`. If you observe
carefully, the result is not the duration in days but an object of class
`interval`. Now let us learn how we can use intervals.

```{r lub10}
interval(course_start, course_end)
```

## Intervals Overlap

Let us say you are planning a vacation and want to check if the vacation
dates overlap with the course dates. You can do this by:

- creating vacation and course intervals
- use `int_overlaps()` to check if two intervals overlap. It returns `TRUE`
if the intervals overlap else `FALSE`.

Let us use the vacation start and end dates to create `vacation_interval`
and then check if it overlaps with `course_interval`. 

```{r lub60, collapse = TRUE}
vacation_start <- as_date('2017-04-19')

vacation_end <- as_date('2017-04-25')

course_interval <- interval(course_start, course_end)

vacation_interval <- interval(vacation_start, vacation_end)

int_overlaps(course_interval, vacation_interval)
```

## How many invoices were settled within due date?

Let us use intervals to count the number of invoices that were settled within
the due date. To do this, we will:

- create an interval for the invoice and due date
- another interval for the due date and the payment date
- if the intervals overlap, the payment was made within the due date 

```{r lub7}
transact %>%
  mutate(
    inv_due_interval = interval(Invoice, Due),
    due_next = Due + days(1),
    due_pay_interval = interval(due_next, Payment),
    overlaps = int_overlaps(inv_due_interval, due_pay_interval)
  ) %>%
  select(Invoice, Due, Payment, overlaps)
```

Below we show another method to count the number of invoices paid within the
due date. Instead of using `days` to change the due date, we use `int_shift`
to shift it by 1 day.

```{r lub12}
transact %>%
  mutate(
    inv_due_interval = interval(Invoice, Due),
    due_pay_interval = interval(Due, Payment),  
    due_pay_next = int_shift(due_pay_interval, by = days(1)),
    overlaps = int_overlaps(inv_due_interval, due_pay_next)
  ) %>%
  select(Invoice, Due, Payment, overlaps)
```


## Shift Interval

Intervals can be shifted too. In the below example, we shift the course 
interval by:

- 1 day
- 3 weeks
- 1 year 

```{r lab50, collapse = TRUE}
course_interval <- interval(course_start, course_end)

course_interval %>%
  int_shift(by = days(1))

course_interval %>%
  int_shift(by = weeks(3))

course_interval %>%
  int_shift(by = years(1))
```

## Within

Does a conference that we need to attend occur during the
course duration?

```{r lub30, collapse = TRUE}
conference <- as_date('2017-04-15')

conference %within% interval(course_start, course_end)
```

#### How many invoices were settled within due date?

Let us use `%within%` to count the number of invoices that were settled within 
the due date. We will do this by:

- creating an interval for the invoice and due date
- check if the payment date falls within the above interval

```{r lub13}
transact %>%
  mutate(
    inv_due_interval = interval(Invoice, Due),
    overlaps = Payment %within% inv_due_interval
  ) %>%
  select(Due, Payment, overlaps)
```

## Quarter

Let us check the quarter and the semester in which the course starts.

```{r lub20, collapse = TRUE}
course_start

course_start %>%
  quarter

course_start %>%
  quarter(with_year = TRUE)

course_start %>%
  semester
```

Let us count the invoices due for each quarter.

```{r lub8}
transact %>%
  mutate(
    quarter_due = quarter(Due)
  ) %>%
  count(quarter_due)
```

```{r lub14}
transact %>%
  mutate(
    Quarter = quarter(Due, with_year = TRUE)
  )
```

## Duration

Let us get the course duration in different units.

```{r lub15, collapse = TRUE}
course_duration <- 
  (course_end - course_start) %>%
  as.numeric

course_duration %>%
  duration(units = 'hours')

course_duration %>%
  duration(units = 'days')

course_duration %>%
  duration(units = 'weeks')
```

#### Case Study

Let us also get the course interval in different units.

```{r lub11, collapse = TRUE}
interval(course_start, course_end) / dseconds()

interval(course_start, course_end) / dminutes()

interval(course_start, course_end) / dhours()

interval(course_start, course_end) / dweeks()

interval(course_start, course_end) / dyears()
```

We can use `time_length()` to get the course interval in different units.

```{r lub16, collapse = TRUE}
interval(course_start, course_end) %>%
  time_length(unit = "seconds")

interval(course_start, course_end) %>%
  time_length(unit = "minutes")

interval(course_start, course_end) %>%
  time_length(unit = "hours")
```

`as.period()` is another way to get the course interval in different units.

```{r lub17, collapse = TRUE}
interval(course_start, course_end) %>%
  as.period(unit = "seconds")

interval(course_start, course_end) %>%
  as.period(unit = "minutes")

interval(course_start, course_end) %>%
  as.period(unit = "hours")
```
