---
title: Hacking strings with stringr
author: Aravind Hebbali
date: '2018-01-07'
slug: hacking-strings-with-stringr
categories:
  - data wrangling
tags:
  - strings
---

## Introduction

In this post, we will learn to work with string data in R using [stringr](http://stringr.tidyverse.org). As we did in the other posts, we will use a case study to explore the various features of the stringr package. Let us begin by installing and loading stringr and a set of other pacakges we will be using.
 
## Libraries, Code & Data

We will use the following libraries: 

- [stringr](http://lubridate.tidyverse.org/index.html) 
- [dplyr](http://dplyr.tidyverse.org/index.html) 
- [magrittr](http://magrittr.tidyverse.org/index.html) 
- [tibble](http://tibble.tidyverse.org/index.html) 
- [purrr](http://purrr.tidyverse.org/index.html) 
- and [readr](http://readr.tidyverse.org/index.html) 

The data sets can be downloaded from [here](https://github.com/rsquaredacademy/datasets) and the codes from [here](https://gist.github.com/aravindhebbali/5d1799744e55dc76cdf1af6b1cc03c82).


```{r str1, message=FALSE}
library(stringr)
library(tibble)
library(magrittr)
library(purrr)
library(dplyr)
library(readr)
```

## Case Study

- extract domain name from random email ids
- extract image type from url
- extract image dimension from url
- extract extension from domain name
- extract http protocol from url
- extract file type from url

### Data

```{r show, message=FALSE, collapse=FALSE}
mockstring <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/mock_strings.csv')
mockstring
```

### Data Dictionary

- domain: dummy website domain
- imageurl: url of an image
- email: dummy email id
- filename: dummy file name with different extensions
- phone: dummy phone number
- address: dummy address with door and street names
- url: randomyly generated urls
- full_name: dummy first and last names
- currency: different currencies
- passwords: dummy passwords

## Overview

Before we start with the case study, let us take a quick tour of **stringr** and introduce
ourselves to some of the functions we will be using later in the case study. One of the 
columns in the case study data is `email`. It contains random email ids. We want to ensure 
that the email ids adher to a particular format .i.e

- they contain `@`
- they contain only one `@`

Let us first detect if the email ids contain `@`. Since the data set has 1000 rows, we will 
use a smaller sample in the examples.

```{r str61, collapse=FALSE}
mockdata <- slice(mockstring, 1:10)
mockdata
```

Use `str_detect()` to detect `@` and `str_count()` to count the number of times
`@` appears in the email ids.

```{r str62, collapse=FALSE}
# detect @
str_detect(mockdata$email, pattern = "@")

# count @
str_count(mockdata$email, pattern = "@")
```

We can use `str_c()` to concatenate strings. Let us add the string `email id:` before each
email id in the data set.

```{r str63}
str_c("email id:", mockdata$email)
```

If we want to split a string into two parts using a particular pattern, we use `str_split()`. 
Let us split the domain name and extension from the domain column in the data. The domain name
and extension are separated by `.` and we will use it to split the domain column. Since `.` is 
a special character, we will use two slashes to escape the special character. 

```{r str64}
str_split(mockdata$domain, pattern = "\\.")
```

We can truncate a string using `str_trunc()`. The default truncation happens at the beggining
of the string but we can truncate the central part or the end of the string as well. 

```{r str65, collapse=FALSE}
str_trunc(mockdata$email, width = 10)
str_trunc(mockdata$email, width = 10, side = "left")
str_trunc(mockdata$email, width = 10, side = "center")
```

Strings can be sorted using `str_sort()`. Let us quickly sort the emails in both
ascending and descending orders.

```{r str66, collapse=FALSE}
str_sort(mockdata$email)
str_sort(mockdata$email, decreasing = TRUE)
```

The case of a string can be changed to upper, lower or title case as shown below.

```{r str67, collapse=FALSE}
str_to_upper(mockdata$full_name)
str_to_lower(mockdata$full_name)
```

Parts of a string can be replaced using `str_replace()`. In the `address` column of the data set,
let us replace:

- Street with ST
- Road with RD

```{r str68, collapse=FALSE}
str_replace(mockdata$address, "Street", "ST")
str_replace(mockdata$address, "Road", "RD")
```

We can extract parts of the string that match a particular pattern using `str_extract()`.

```{r str69}
str_extract(mockdata$email, pattern = "org")
```

Before we extract, we need to know whether the string contains text that match our pattern.
Use `str_match()` to see if the pattern is present in the string.

```{r str70}
str_match(mockdata$email, pattern = "org")
```

If we are dealing with a character vector and know that the pattern we are looking at
is present in the vector, we might want to know the index of the strings in which it is
present. Use `str_which()` to identify the index of the strings that match our pattern.

```{r str71}
str_which(mockdata$email, pattern = "org")
```

Another objective might be to locate the position of the pattern we are looking for in the 
string. For example, if we want to know the position of `@`  in the email ids, we can use
`str_locate()`.

```{r str72}
str_locate(mockdata$email, pattern = "@")
```

The length of the string can be computed using `str_length()`. Let us ensure that the length 
of the strings in the `password` column is 16.

```{r str74}
str_length(mockdata$passwords)
```

We can extract parts of a string by specifying the starting and ending position using
`str_sub()`. Let us extract the currency type from the `currency` column.

```{r str75}
str_sub(mockdata$currency, start = 1, end = 1)
```

One final function that we will look at before the case study is `word()`. It extracts 
word(s) from sentences. We do not have any senetences in the data set, but let us use it
to extract the first and last name from the `full_name` column.

```{r 73, collapse=FALSE}
word(mockdata$full_name, 1)
word(mockdata$full_name, 2)
```

Alright, now let us apply what we have learned so far to our case study.

## Extract domain name from email ids

### Steps

- split email using pattern `@`
- extract the second element from the resulting list
- split the above using pattern `\\.`
- extract the first element from the resulting list

Let us take a look at the emails before we extract the domain names.

```{r str15}
emails <- 
  mockstring %>%
  pull(email) %>%
  head

emails
```

#### Step 1: Split email using pattern `@`.

We will split the email using `str_split`. It will split a string 
using the pattern supplied. In our case the pattern is `@`.

```{r str16}
 str_split(emails, pattern = '@')
```

#### Step 2: Extract the second element from the resulting list.

Step 1 returned a list. Each element of the list has two values. The first 
one is the username and the second is the domain name. Since we are 
extracting the domain name, we want the second value from each element of 
the list.

We will use `map_chr()` from purrr to extract the domain names. It will
return the second value from each element in the list. Since the domain 
name is a string, `map_chr()` will return a character vector.

```{r str17}
emails %>%
  str_split(pattern = '@') %>%
  map_chr(2)
```

#### Step 3: Split the above using pattern `\\.`.

We want the domain name and not the extension. Step 2 returned a 
character vector and we need to split the domain name and the domain
extension. They are separated by `.`. Since `.` is a special character,
we will use `\\` before `.` to escape it. Let us split the domain
name and domain extension using `str_split` and `\\.` as the pattern.

```{r str18}
emails %>%
  str_split(pattern = '@') %>%
  map_chr(2) %>%
  str_split(pattern = '\\.') 
```

#### Step 4: Extract the first element from the resulting list.

Now that we have separated the domain name from its extension, let us extract
the first value from each element in the list returned in step 3. We will again 
use `map_chr` to achieve this.

```{r str19}
emails %>%
  str_split(pattern = '@') %>%
  map_chr(2) %>%
  str_split(pattern = '\\.') %>%
  map_chr(extract(1))
```

## Extract Domain Extension

The below code extracts the domain extension instead of the domain name. 

```{r str20}
emails %>%
  str_split(pattern = '@') %>%
  map_chr(2) %>%
  str_split(pattern = '\\.', simplify = TRUE) %>%
  extract(, 2)
```

## Extract image type from URL

### Steps

- split imageurl using pattern `\\.`
- extract the third value from each element of the resulting list
- subset the string using the index position 

Let us take a look at the URL of the image.

```{r str21}
img <- 
  mockstring %>%
  pull(imageurl) %>%
  head

img
```

#### Step 1: Split imageurl using pattern `\\.`

Let us split imageurl using `str_split` and the pattern `\\.`.

```{r str22}
img %>%
  str_split(pattern = '\\.')
```

#### Step 2: Extract the third value from each element of the resulting list

Step 1 returned a list the elements of which have 3 values each. If you
observe the list, the image type is in the 3rd value. We will now 
extract the third value from each element of the list using `map_chr`.

```{r str23}
img %>%
  str_split(pattern = '\\.') %>%
  map_chr(extract(3))
```

#### Step 3: Subset the string using the index position 

We can now extract the image type in two ways:

- subset the first 3 characters of the string
- split the string using pattern `/` and extract the first value from the 
elements of the resulting list

Below is the first method. We know that the image type is 3 characters. So 
we use `str_sub` to subset the first 3 characters. The index positions 
are mentioned using `start` and `stop`.

```{r str24}  
img %>%
  str_split(pattern = '\\.') %>%
  map_chr(extract(3)) %>%
  str_sub(start = 1, end = 3)
``` 

In case you are not sure about the length of the image type. In such cases,
we will split the string using pattern `/` and then use `map_chr` to
extract the first value of each element of the resulting list.

```{r str25}
img %>%
  str_split(pattern = '\\.') %>%
  map_chr(extract(3)) %>%
  str_split(pattern = '/') %>%
  map_chr(extract(1))
```

## Extract Image Dimesion from URL

### Steps

- locate numbers between 0 and 9
- extract part of url starting with image dimension
- split the string using the pattern `\\.`
- extract the first element

#### Step 1: Locate numbers between 0 and 9.

Let us inspect the image url. The dimension of the image appears
after the domain extension and there are no numbers in the url
before. We will locate the position or index of the first number 
in the url using `str_locate()` and using the pattern `[0-9]` 
which instructs to look for any number between and including 0 
and 9.

```{r str26}
img %>%
  str_locate(pattern = "[0-9]") 
```

#### Step 2: Extract url 

We know where the dimension is located in the url. Let us extract the
part of the url that contains the image dimension using `str_sub()`.

```{r str27}
img %>%
  str_sub(start = 23) 
```

#### Step 3: Split the string using the pattern `\\.`.

From the previous step, we have the part of the url that
contains the image dimension. To extract the dimension, we
will split it from the rest of the url using `str_split()`
and using the pattern `\\.` as it separates the dimension
and the image extension.

```{r str28}
img %>%
  str_sub(start = 23) %>%
  str_split(pattern = '\\.') 
```

#### Step 4: Extract the first element.

The above step resulted in a list which contains the 
image dimension and the rest of the url. Each element
of the list is a character vector. We want to extract 
the first value in the character vector. Let us use
`map_chr()` to extract the first value from each
element of the list.

```{r str29}
img %>%
  str_sub(start = 23) %>%
  str_split(pattern = '\\.') %>%
  map_chr(extract(1))
```

## Extract HTTP Protocol from URL

```{r str30}
url1 <- 
  mockstring %>%
  pull(url) %>%
  first

url1
```

### Steps

- split the url using the pattern `://`
- extract the first element

#### Step 1: Split the url using the pattern `://`.

The HTTP protocol is the first part of the url and is
separated from the rest of the url by `:`. Let us
split the url using `str_split()` and using the
pattern `:`. Since `:` is a special character, we
will escape it using `\\`.

```{r str31}
url1 %>%
  str_split(pattern = '://') 
```

#### Step 2: Extract the first element.

The HTTP protocol is the first value in each element
of the list. As we did in the previous example, we 
will extact it using `map_chr()` and `extract()`.

```{r str32}
url1 %>%
  str_split(pattern = '://') %>%
  map_chr(extract(1))
```


## Extract file type

```{r str14}
urls <-
  mockstring %>%
  use_series(url) %>%
  extract(1:3)
```

### Steps

- check if there are only 2 dots in the URL
- check if there is only 1 question mark in the URL
- detect the staritng position of file type
- tetect the ending position of file type
- use the locations to specify the index position for extracting file type

#### Step 1: Check if there are only 2 dots in the URL

Let us locate all the dots in the url using `str_locate_all()` and see 
if any of them contain more than 2 dots.

```{r str9}
urls %>%
  str_locate_all(pattern = '\\.') %>%
  map_int(nrow) %>%
  is_greater_than(2) %>%
  sum
```

#### Step 2: Check if there is only 1 question mark in the URL

The next step is to check if there is only one `?` (question mark)
in the url. 

```{r str10}  
urls %>%
  str_locate_all(pattern = "[?]") %>%
  map_int(nrow) %>%
  is_greater_than(1) %>%
  sum
```

#### Step 3: Detect the staritng position of file type

Since the file type is located between the second dot and 
the first quesiton mark in the url, let us extract the
location of the second dot and add 1 as the file type 
starts after the dot.

```{r str11}
d <- 
  urls %>%
  str_locate_all(pattern = '\\.') %>%
  map_int(extract(2)) %>%
  add(1)

d  
```

#### Step 4: Detect the ending position of file type

In step 2, we confirmed that the url has only one question mark. 
Let us locate the question mark in the url and subtract 1 (as the 
file type ends before the question mark) so that we get the 
ending postion of the file type.
.


```{r str12}
q <-  
  urls %>%
  str_locate_all(pattern = "[?]") %>%
  map_int(extract(1)) %>%
  subtract(1)

q
```

#### Step 5: Specify the index position for extracting file type

From steps 3 and 4, we have the location of the second dot and the 
first question mark in the url. Let us use them with `str_sub()` 
to extract the file type.

```{r str13}
urls %>% 
  str_sub(start = d, end = q)
```