---
title: 'ggplot2: Box Plots'
author: Aravind Hebbali
twitterImg: /img/gg_box.png
description: "ggplot2 - Box Plots"
date: '2018-01-07'
slug: ggplot2-box-plots
topics:
  - ggplot2
tags:
  - ggplot2
  - visualization
---

### Introduction

This is the 9th post in the series **Elegant Data Visualization with 
ggplot2**. In the previous post, we learnt how to build bar charts. In this 
post, we will learn to:

- build box plots
- modify box 
    - color
    - fill
    - alpha
    - line size
    - line type
- modify outlier 
    - color
    - shape 
    - size 
    - alpha

The box plot is a standardized way of displaying the distribution of data. It 
is useful for detecting outliers and for comparing distributions and shows the 
shape, central tendancy and variability of the data.

<br>

## Structure

- the body of the boxplot consists of a "box" (hence, the name), which goes from the first quartile (Q1) to the third quartile (Q3)
- within the box, a vertical line is drawn at the Q2, the median of the data set
- two horizontal lines, called whiskers, extend from the front and back of the box 
- the front whisker goes from Q1 to the smallest non-outlier in the data set, and the back whisker goes from Q3 to the largest non-outlier
- if the data set includes one or more outliers, they are plotted separately as points on the chart


```{r box, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

<br>

### Libraries, Code & Data

We will use the following libraries in this post:

- [readr](http://readr.tidyverse.org/)
- [ggplot2](http://ggplot2.tidyverse.org/)

All the data sets used in this post can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/246091b512a6c006e68374e2d24caf7c).

<br>

## Data

We are going to use two different data sets in this post. Both the data sets have the same data but are in 
different formats. 

```{r show, eval=TRUE, message=FALSE}
daily_returns <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/tickers.csv')
daily_returns
```

<br>

## Univariate Box Plot

If you are not comparing the distribution of continuous data, you can create 
box plot for a single variable. Unlike `plot()`, where we could just use 
1 input, in ggplot2, we must specify a value for the X axis and it must be
categorical data. Since we are not comparing distributions, we will use `1`
as the value for the X axis and wrap it inside `factor()` to treat it as a
categorical variable. In the below example, we examine the distribution of 
stock returns of Apple. 

```{r box2, fig.align='center'}
ggplot(daily_returns) +
  geom_boxplot(aes(x = factor(1), y = AAPL))
```

<br>

## Data

For the rest of the post, we will use the below data set. Instead of 5 columns, 
we have two columns. One for the stock names and another for returns. 

```{r show2, eval=TRUE, message=FALSE}
tidy_returns <- 
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/tidy_tickers.csv',
  col_types = list(col_factor(levels = c('AAPL', 'AMZN', 'FB', 'GOOG', 'MSFT')), col_double()))
tidy_returns
```

<br>

## Box Plot

With the above data, let us create a box plot where we compate the distribution
of stock returns of different companies. We map X axis to the column with stock 
names and Y axis to the column with stock returns. Note that, the column names 
are wrapped inside `aes()`.

```{r box3, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns))
```

<br>

To create a horizontal bar plot, we can use `coord_flip()` which will flip the
coordinate axes.

## Horizontal Box Plot

```{r box4, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns)) +
  coord_flip()
```

<br>

## Notch

Notches are used to compare medians. You can use the `notch` argument and set
it to `TRUE`. 

```{r box5, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns),
    notch = TRUE) 
```

<br>

## Jitter

Just for comparison, let us plot the returns as points on top of the box plot
using `geom_jitter()`. We modify the color of the points using the `color`
argument and the spread using the `width` argument.

```{r box6, fig.align='center'}
ggplot(tidy_returns, aes(x = stock, y = returns)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, color = 'blue')
```

<br>

## Outliers

To highlight extreme observations, we can modify the appearance of outliers 
using the following:

- color
- shape
- size
- alpha

<br>

To modify the color of the outliers, use the `outlier.color` argument. The 
color can be specified either using its name or the associated hex code.

```{r box7, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), outlier.color = 'red')
```

<br>

The shape of the outlier can be modified using the `outlier.shape` argument.
It can take values between `0` and `25`.

```{r box9, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), outlier.shape = 23) 
```

<br>

The size of the outlier can be modified using the `outlier.size` argument. It 
can take any value greater than `0`.

```{r box10, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), outlier.size = 3) 
```

<br>

You can play around with the transparency of the outlier using the 
`outlier.alpha` argument. It can take values between `0` and `1`.

```{r box11, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), outlier.color = 'blue', outlier.alpha = 0.3) 
```

<br>

## Box Aesthetics

The appearance of the box can be controlled using the following:

- color
- fill
- alpha
- line type
- line width

<br>

## Specify Values 

The background color of the box can be modified using the `fill` argument. The 
color can be specified either using its name or the associated hex code.

```{r box12, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), fill = c('blue', 'red', 'green', 'yellow', 'brown')) 
```

<br>

To modify the transparency of the background color, use the `alpha` argument. It 
can take any value between `0` and `1`.

```{r box14, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), fill = 'blue', alpha = 0.3) 
```

<br>

The color of the border can be modified using the `color` argument. The 
color can be specified either using its name or the associated hex code.

```{r box15, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), color = c('blue', 'red', 'green', 'yellow', 'brown')) 
```

<br>

The width of the border can be changed using the `size` argument. It can take
any value greater than `0`.

```{r box17, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), size = 1.5) 
```

<br>

To change the line type of the border, use the `linetype` argument. It can take
any value between `0` and `6`.

```{r box18, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns), linetype = 2) 
```

<br>

## Map Variables

Instead of specifying values, we can map `fill` and `color` to variables as 
well. In the below example, we map `fill` to the variable stock. It assigns
different colors to the different stocks.

```{r box13, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns, fill = stock)) 
```

<br>

Let us map `color` to the variable stock. It will assign different colors
to the box borders.

```{r box16, fig.align='center'}
ggplot(tidy_returns) +
  geom_boxplot(aes(x = stock, y = returns, color = stock)) 
```

<br>

### Summary

In this post, we learnt to:

- build box plots
- modify outlier color, shape, size etc.
- modify box color
- modify box line color, size and line type

<br>

### Up Next..

In the next post, we will learn to build histograms.

<br>