---
title: Introducing olsrr
author: Aravind Hebbali
twitterImg: /img/og_olsrr.png
description: "Tools for linear regression."
date: '2019-02-08'
slug: introducing-olsrr
topics:
  - regression
tags:
  - olsrr
---

I am pleased to announce the **olsrr** package, a set of tools for improved 
output from linear regression models, designed keeping in mind 
beginner/intermediate R users. The package includes:

- comprehensive regression output
- variable selection procedures
- heteroskedasticiy, collinearity diagnostics and measures of influence
- various plots and underlying data

If you know how to build models using `lm()`, you will find **olsrr** very 
useful. Most of the functions use an object of class `lm` as input. So you 
just need to build a model using `lm()` and then pass it onto the functions in 
**olsrr**. Once you have picked up enough knowledge of R, you can move on to 
more intuitive approach offered by tidymodels etc. as they offer more 
flexibility, which **olsrr** does not.

### Installation

```{r cran-installation, eval = FALSE}
# Install release version from CRAN
install.packages("olsrr")

# Install development version from GitHub
# install.packages("devtools")
devtools::install_github("rsquaredacademy/olsrr")
```

### Shiny App

**olsrr** includes a shiny app which can be launched using

```{r app, eval=FALSE}
ols_launch_app()
```

or try the live version [here](https://www.rsquaredcomputing.com/olsrr/).

Read on to learn more about the features of **olsrr**, or see the 
[olsrr website](https://olsrr.rsquaredacademy.com/) for 
detailed documentation on using the package.

### Regression Output

```{r load, message=FALSE, echo=FALSE}
library(olsrr)
```

```{r regress}
model <- lm(mpg ~ disp + hp + wt + qsec, data = mtcars)
ols_regress(model)
```

In the presence of interaction terms in the model, the predictors are scaled 
and centered before computing the standardized betas. `ols_regress()` will 
detect interaction terms automatically but in case you have created a new 
variable instead of using the inline function, you can indicate the presence 
of interaction terms by setting `iterm` to `TRUE`.

### Residual Diagnostics

**olsrr** offers tools for detecting violation of standard regression assumptions:

- Residual QQ plot
- Residual normality test
- Residual vs Fitted plot
- Residual histogram

```{r residiag}
ols_plot_resid_qq(model)
```

See [Residual Diagnostics](https://olsrr.rsquaredacademy.com/articles/residual_diagnostics.html) 
for more details.

### Heteroskedasticity

**olsrr** provides the following 4 tests for detecting heteroscedasticity:

- Bartlett Test
- Breusch Pagan Test
- Score Test
- F Test

```{r hetero}
ols_test_breusch_pagan(model)
```

See [Heteroskedasticity](https://olsrr.rsquaredacademy.com/articles/heteroskedasticity.html) 
for more details.

### Collinearity Diagnostics

VIF, Tolerance and condition indices to detect collinearity and plots for 
assessing mode fit and contributions of variables.

```{r collinearity}
ols_coll_diag(model)
```

See [Collinearity Diagnostics](https://olsrr.rsquaredacademy.com/articles/regression_diagnostics.html) for more details.

### Measures of Influence

**olsrr** offers the following tools to detect influential observations:

- Cook’s D Bar Plot
- Cook’s D Chart
- DFBETAs Panel
- DFFITs Plot
- Studentized Residual Plot
- Standardized Residual Chart
- Studentized Residuals vs Leverage Plot
- Deleted Studentized Residual vs Fitted Values Plot
- Hadi Plot
- Potential Residual Plot

```{r mof}
ols_plot_resid_lev(model)
```

See [Measures of Influence](https://olsrr.rsquaredacademy.com/articles/influence_measures.html) for more details.

### Variable Selection

Different variable selection procedures such as all possible regression, best 
subset regression, stepwise regression, stepwise forward regression and 
stepwise backward regression.

```{r varselect}
model <- lm(y ~ ., data = stepdata)
ols_step_both_aic(model)
```

See [Variable Selection](https://olsrr.rsquaredacademy.com/articles/variable_selection.html) for more details.

### Learning More

The [olsrr website](https://olsrr.rsquaredacademy.com/index.html) includes 
comprehensive documentation on using the package, including the following
articles that cover various aspects of using olsrr:

- [Variable Selection](https://olsrr.rsquaredacademy.com/articles/variable_selection.html) - Different variable selection procedures such as all possible regression, best 
subset regression, stepwise regression, stepwise forward regression and 
stepwise backward regression.

- [Residual Diagnostics](https://olsrr.rsquaredacademy.com/articles/residual_diagnostics.html) - Includes plots to examine residuals to validate OLS assumptions.

- [Heteroskedasticity](https://olsrr.rsquaredacademy.com/articles/heteroskedasticity.html) - Tests for heteroskedasticity include bartlett test, breusch pagan test, score test and f test.

- [Collinearity Diagnostics](https://olsrr.rsquaredacademy.com/articles/regression_diagnostics.html) - VIF, Tolerance and condition indices to detect collinearity and plots for assessing mode fit and contributions of variables.

- [Measures of Influence](https://olsrr.rsquaredacademy.com/articles/influence_measures.html) - Includes 10 different plots to detect and identify influential observations.


### Feedback

**olsrr** has been on CRAN for more than an year while we were fixing bugs and 
making the API stable. All feedback is welcome. Issues (bugs and feature 
requests) can be posted to [github tracker](https://github.com/rsquaredacademy/olsrr/issues).
For help with code or other related questions, feel free to reach me hebbali.aravind@gmail.com.


