---
title: Descriptive/Summary Statistics with descriptr
author: Aravind Hebbali
twitterImg: /img/og_descriptr.png
description: "Tools for descriptive statistics in R."
date: '2019-02-20'
slug: introducing-descriptr
topics:
  - summary statistics
tags:
  - descriptr
---

We are pleased to introduce the **descriptr** package, a set of tools for 
generating descriptive/summary statistics.

## Installation

```{r cran-installation, eval = FALSE}
# Install release version from CRAN
install.packages("descriptr")

# Install development version from GitHub
# install.packages("devtools")
devtools::install_github("rsquaredacademy/descriptr")
```

## Shiny App

**descriptr** includes a shiny app which can be launched using

```{r app, eval=FALSE}
ds_launch_shiny_app()
```

or try the live version [here](https://www.rsquaredcomputing.com/descriptr/).

Read on to learn more about the features of **descriptr**, or see the 
[descriptr website](https://descriptr.rsquaredacademy.com/) for 
detailed documentation on using the package.

```{r, echo=FALSE, message=FALSE}
library(descriptr)
library(dplyr)
```

## Data

We have modified the `mtcars` data to create a new data set `mtcarz`. The only
difference between the two data sets is related to the variable types. 

```{r egdata}
str(mtcarz)
```

## Data Screening

The `ds_screener()` function will screen a data set and return the following:
- Column/Variable Names
- Data Type
- Levels (in case of categorical data)
- Number of missing observations
- % of missing observations

```{r screener}
ds_screener(mtcarz)
```

## Continuous Data

### Summary Statistics

The `ds_summary_stats()` function returns a comprehensive set of statistics 
including measures of location, variation, symmetry and extreme observations.

```{r summary}
ds_summary_stats(mtcarz, mpg)
```

You can pass multiple variables as shown below:

```{r summary2}
ds_summary_stats(mtcarz, mpg, disp)
```

If you do not specify any variables, it will detect all the continuous 
variables in the data set and return summary statistics for each of them.

### Frequency Distribution

The `ds_freq_table()` function creates frequency tables for continuous variables. 
The default number of intervals is 5.

```{r fcont}
ds_freq_table(mtcarz, mpg, 4)
```

#### Histogram

A `plot()` method has been defined which will generate a histogram.

```{r fcont_hist, fig.width=7, fig.height=7, fig.align='centre'}
k <- ds_freq_table(mtcarz, mpg, 4)
plot(k)
```

### Auto Summary

If you want to view summary statistics and frequency tables of all or subset of
variables in a data set, use `ds_auto_summary()`.

```{r auto-summary}
ds_auto_summary_stats(mtcarz, disp, mpg)
```

### Group Summary

The `ds_group_summary()` function returns descriptive statistics of a continuous
variable for the different levels of a categorical variable.

```{r gsummary}
k <- ds_group_summary(mtcarz, cyl, mpg)
k
```

`ds_group_summary()` returns a tibble which can be used for further analysis.

```{r gsummary_tibble}
k$tidy_stats
```

#### Box Plot

A `plot()` method has been defined for comparing distributions.

```{r gsum_boxplot, fig.width=7, fig.height=7, fig.align='centre'}
k <- ds_group_summary(mtcarz, cyl, mpg)
plot(k)
```

### Multiple Variables

If you want grouped summary statistics for multiple variables in a data set, use
`ds_auto_group_summary()`.

```{r auto-group-summary}
ds_auto_group_summary(mtcarz, cyl, gear, mpg)
```

## Multiple Variable Statistics 

The `ds_tidy_stats()` function returns summary/descriptive statistics for 
variables in a data frame/tibble.

```{r multistats}
ds_tidy_stats(mtcarz, mpg, disp, hp)
```

### Measures

If you want to view the measure of location, variation, symmetry, percentiles 
and extreme observations as tibbles, use the below functions. All of them, 
except for `ds_extreme_obs()` will work with single or multiple variables. If 
you do not specify the variables, they will return the results for all the 
continuous variables in the data set.

#### Measures of Location

```{r mloc}
ds_measures_location(mtcarz)
```

#### Measures of Variation

```{r mvar}
ds_measures_variation(mtcarz)
```

#### Measures of Symmetry

```{r msym}
ds_measures_symmetry(mtcarz)
```

#### Percentiles

```{r mperc}
ds_percentiles(mtcarz)
```

## Categorical Data

### Cross Tabulation

The `ds_cross_table()` function creates two way tables of categorical variables. 

```{r cross}
ds_cross_table(mtcarz, cyl, gear)
```

If you want the above result as a tibble, use `ds_twoway_table()`.

```{r cross_tibble}
ds_twoway_table(mtcarz, cyl, gear)
```

A `plot()` method has been defined which will generate:

#### Grouped Bar Plots

```{r cross_group, fig.width=7, fig.height=7, fig.align='centre'}
k <- ds_cross_table(mtcarz, cyl, gear)
plot(k)
```

#### Stacked Bar Plots

```{r cross_stack, fig.width=7, fig.height=7, fig.align='centre'}
k <- ds_cross_table(mtcarz, cyl, gear)
plot(k, stacked = TRUE)
```

#### Proportional Bar Plots

```{r cross_prop, fig.width=7, fig.height=7, fig.align='centre'}
k <- ds_cross_table(mtcarz, cyl, gear)
plot(k, proportional = TRUE)
```

### Frequency Table 

The `ds_freq_table()` function creates frequency tables.

```{r ftable}
ds_freq_table(mtcarz, cyl)
```

A `plot()` method has been defined which will create a bar plot.

```{r ftable_bar, fig.width=7, fig.height=7, fig.align='centre'}
k <- ds_freq_table(mtcarz, cyl)
plot(k)
```

### Multiple One Way Tables

The `ds_auto_freq_table()` function creates multiple one way tables by creating a 
frequency table for each categorical variable in a data set. You can also 
specify a subset of variables if you do not want all the variables in the data 
set to be used.

```{r oway}
ds_auto_freq_table(mtcarz)
```

### Multiple Two Way Tables

The `ds_auto_cross_table()` function creates multiple two way tables by creating a 
cross table for each unique pair of categorical variables in a data set. You 
can also specify a subset of variables if you do not want all the variables in 
the data set to be used.

```{r tway}
ds_auto_cross_table(mtcarz, cyl, gear, am)
```

## Visualization

**descriptr** can help visualize multiple variables by automatically
detecting their data types. 

### Continuous Data

```{r scatter}
ds_plot_scatter(mtcarz, mpg, disp, hp)
```

### Categorical Data

```{r bar_stacked}
ds_plot_bar_stacked(mtcarz, cyl, gear, am)
```

## Learning More

The [descriptr website](https://descriptr.rsquaredacademy.com/index.html) includes 
comprehensive documentation on using the package, including the following
articles that cover various aspects of using rfm:

- [Continuous Data](https://descriptr.rsquaredacademy.com/articles/continuous-data.html) - for 
summarizing continuous data.

- [Categorical Data](https://descriptr.rsquaredacademy.com/articles/categorical-data.html) - for 
summarizing categorical data.

- [Visualization](https://descriptr.rsquaredacademy.com/articles/visualization.html) - for 
generating different types of plots.

## Feedback

All feedback is welcome. Issues (bugs and feature 
requests) can be posted to [github tracker](https://github.com/rsquaredacademy/descriptr/issues).
For help with code or other related questions, feel free to reach me hebbali.aravind@gmail.com.


