---
title: Binning Data with rbin 
author: Aravind Hebbali
description: "Binning Data with rbin (Includes RStudio Addins)"
date: '2019-03-12'
slug: binning-data-with-rbin
categories:
  - binning
tags:
  - rbin
---

We are happy to introduce the **rbin** package, a set of tools for binning/discretization
of data, designed keeping in mind beginner/intermediate R users. It comes with
two RStudio addins for interactive binning.

### Installation

```{r cran-installation, eval = FALSE}
# Install release version from CRAN
install.packages("rbin")

# Install development version from GitHub
# install.packages("devtools")
devtools::install_github("rsquaredacademy/rbin")
```

### RStudio Addins

**rbin** includes two RStudio addins for manually binning data. Below
is a demo:

```{r tube1, eval=TRUE, echo=FALSE}
blogdown::shortcode("youtube", "KQGPB1sL7R8")
```

Read on to learn more about the features of **rbin**, or see the 
[rbin website](https://rbin.rsquaredacademy.com/) for 
detailed documentation on using the package.

### Introduction

```{r, echo=FALSE, message=FALSE}
library(rbin)
options(knitr.table.format = "html")
options(tibble.width = Inf)
```

Binning is the process of transforming numerical or continuous data into 
categorical data. It is a common data pre-processing step of the model building 
process. **rbin** has the following features:

- manual binning using shiny app
- equal length binning method
- winsorized binning method
- quantile binning method
- combine levels of categorical data
- create dummy variables based on binning method
- calculates weight of evidence (WOE), entropy and information value (IV)
- provides summary information about binning process

### Manual Binning

For manual binning, you need to specify the cut points for the bins. **rbin** 
follows the left closed and right open interval (`[0,1) = {x | 0 ≤ x < 1}`) for 
creating bins. The number of cut points you specify is one less than the number 
of bins you want to create i.e. if you want to create 10 bins, you need to 
specify only 9 cut points as shown in the below example. The accompanying 
RStudio addin, `rbinAddin()` can be used to iteratively bin the data and to 
enforce monotonic increasing/decreasing trend.

After finalizing the bins, you can use `rbin_create()` to create the dummy 
variables.

#### Bins

```{r manual}
bins <- rbin_manual(mbank, y, age, c(29, 31, 34, 36, 39, 42, 46, 51, 56))
bins
```

#### Plot

```{r manual_plot, fig.width=7, fig.height=5, fig.align='center'}
plot(bins)
```

#### Dummy Variables

```{r manual_dummy}
bins <- rbin_manual(mbank, y, age, c(29, 31, 34, 36, 39, 42, 46, 51, 56))
rbin_create(mbank, age, bins)
```

### Factor Binning

You can collapse or combine levels of a factor/categorical variable using 
`rbin_factor_combine()` and then use `rbin_factor()` to look at weight of 
evidence, entropy and information value. After finalizing the bins, you can 
use `rbin_factor_create()` to create the dummy variables. You can use the 
RStudio addin, `rbinFactorAddin()` to interactively combine the levels and 
create dummy variables after finalizing the bins.

#### Combine Levels

```{r factor_combine}
upper <- c("secondary", "tertiary")
out <- rbin_factor_combine(mbank, education, upper, "upper")
table(out$education)

out <- rbin_factor_combine(mbank, education, c("secondary", "tertiary"), "upper")
table(out$education)
```

#### Bins

```{r factor_bins}
bins <- rbin_factor(mbank, y, education)
bins
```

#### Plot

```{r factor_plot, fig.width=7, fig.height=5, fig.align='center'}
plot(bins)
```

#### Create Bins

```{r factor_create}
upper <- c("secondary", "tertiary")
out <- rbin_factor_combine(mbank, education, upper, "upper")
rbin_factor_create(out, education)
```

### Quantile Binning

Quantile binning aims to bin the data into roughly equal groups using quantiles.

```{r quantile}
bins <- rbin_quantiles(mbank, y, age, 10)
bins
```

#### Plot

```{r quantile_plot, fig.width=7, fig.height=5, fig.align='center'}
plot(bins)
```


### Winsorized Binning

Winsorized binning is similar to equal length binning except that both tails 
are cut off to obtain a smooth binning result. This technique is often used 
to remove outliers during the data pre-processing stage. For Winsorized 
binning, the Winsorized statistics are computed first. After the minimum and 
maximum have been found, the split points are calculated the same way as in 
equal length binning. 

```{r winsorize}
bins <- rbin_winsorize(mbank, y, age, 10, winsor_rate = 0.05)
bins
```

#### Plot

```{r winsorize_plot, fig.width=7, fig.height=5, fig.align='center'}
plot(bins)
```

### Learning More

The [rbin website](https://rbin.rsquaredacademy.com/) includes 
comprehensive documentation on using the package, including the following
article that gives a brief introduction to rbin:

- [Introduction to rbin](https://rbin.rsquaredacademy.com/articles/introduction.html) 

### Feedback

**rbin** has been on CRAN for a few months now while we were fixing bugs and 
making the API stable. All feedback is welcome. Issues (bugs and feature 
requests) can be posted to [github tracker](https://github.com/rsquaredacademy/rbin/issues).
For help with code or other related questions, feel free to reach out to us
at pkgs@rsquaredacademy.com.


